### 2025-02-11 paper/main.tex
commit first working paper draft

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%    Canadian AI Latex Template    %
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\documentclass[10pt]{cai}

\begin{document}
% Editorial staff will replace the following values:
% 1. Conference Year
% 2. Issue number
% 3. Article DOI
\def\conferenceyear{2025}
\volumeheader{38}{0}%{00.000}
\begin{center}

\title{Development of a Reinforcement Learning Enabled Cattle Tracker Prototype}
\maketitle

\thispagestyle{empty}

% Add Authors and Affiliations in the camera ready
% for the double blind review, please leave this section as is 
\begin{tabular}{cc}
First Author\upstairs{\affilone,*}, Second Author\upstairs{\affilone}, Third Author\upstairs{\affilthree}
\\[0.25ex]
{\small \upstairs{\affilone} Affiliation One} \\
{\small \upstairs{\affiltwo} Affiliation Two} \\
{\small \upstairs{\affilthree} Affiliation Three} \\
\end{tabular}

% Replace with corresponding author email address
\emails{
  \upstairs{*}corresponding\_author@example.ca 
}
\vspace*{0.2in}
\end{center}

\begin{abstract}
This is just a sample \LaTeX template. As usual, most of the customization for the proceedings is done in the \texttt{cai.cls} file. 
\end{abstract}

% add your keywords
\begin{keywords}{Keywords:}
Provide up to six keywords, separated by commas.
\end{keywords}
\copyrightnotice

\section{About CAIAC}

The Canadian Artificial Intelligence Association / Association Pour L'Intelligence Artificielle au Canada started out of a workshop in the Spring of 1973 and was originally named as the Canadian Society for Computational Studies of Intelligence/Societé canadienne pour études d'intelligence par ordinateur (CSCSI/SCEIO) and is probably the oldest organized association of its kind. The first formal CSCSI/SCEIO conference happened in 1976. The conference was held biannually for several years until it became an yearly event in 2000. In 2008, CSCSI/SCEIO was officially renamed to CAIAC.

More about our history can be found at \url{https://www.caiac.ca/en/history-of-caiac}.

\section{Submission Details}
\label{submission}
We invite submissions of both long and short papers. Long papers must be no longer than \textbf{12 pages}, and Short papers must be no longer than \textbf{6 pages}, including references, formatted using the conference template.

Papers submitted to the conference must contain original work that has not already been published, accepted for publication, or under review by a journal or another conference. Submissions will go through a \textbf{double-blind} review process. Authors are responsible for making sure submissions are anonymized.

A ``Best Paper Award'' and a ``Best Student Paper Award'' will be given at the conference respectively to the authors of each best paper, as judged by the Best Paper Award Selection Committee.

\section{Publication and Registration}
\label{pub}
Since 2020, the conference proceedings are published through PubPub open access\footnote{\url{https://www.pubpub.org/}} and indexed DBLP, ACM, Google Scholar, and other services that access Pubpub publications.

At least one author of each accepted paper is required to attend the conference to present the work. The authors must agree to this requirement prior to submitting their paper for review.

In addition, the corresponding author of each paper, acting on behalf of all of the authors of that paper, must complete and sign a Consent-to-Publish form. The corresponding author signing the copyright form should match the corresponding author marked on the paper.

\section{Figures and Tables}

This simple \LaTeX\xspace template has been tested with \texttt{latex} and with \texttt{pdflatex} commands, directly from the command line or through \texttt{latexmk}. We ask that you place captions below tables and figures as in the examples below.

Pubpub uses a fairly complete \LaTeX\xspace engine that is likely to handle your source files without problems. Please check \url{https://help.pubpub.org/pub/latex-compatibility} if you require specialized packages or non-standard customization.

\begin{figure}[ht]
  \centering
  \ifpdf
    \includegraphics[scale=0.6]{figs/sample_fig.png}
  \else
    \includegraphics[scale=0.6,natwidth=330,natheight=120]{figs/sample_fig.png}
  \fi
  \caption{Captions under figures, please.}
  \label{fig:sample_fig}
\end{figure}

\begin{table}[h]
\begin{tabular}{|l|l|}
\textbf{Header one} & \textbf{Header two} \\ 
\hline
 Some text & More text
\end{tabular}
\vspace{0.2cm}
\caption{Captions under tables, please.}
\label{tab:important}
\end{table}

\section{Using Overleaf}

If you are reading this you have already downloaded the zip file with the template (\textbf{CANAI \conferenceyear\xspace LaTex Template.zip}). All you need to do is to upload it into Overleaf:

\begin{enumerate}
\item Go to \url{overleaf.com}.
\item Click on new project (top left menu).
\item Choose the \textbf{upload project} option.
\item Upload the \textbf{zip file} to Overleaf.
\end{enumerate}

\section*{Acknowledgements}
This section can be left blank during double-blind review.

%Appendixes go here
\appendix

\section{Example of math equation }
%\label{appendix-customize-this-label}
Binomial theorem: \cite{abramowitz1948handbook}
\begin{equation}
(x+y)^n=\sum_{\substack{k=0}}^{n}\dbinom{n}{k}x^{n-k}y^k
\end{equation}

% All references should be stored in the file "references.bib".
% That call to use that file is in "cai.cls". 
% Please do not modify anything below this line.
\printbibliography[heading=subbibintoc]

\end{document}

### 2025-02-12 paper/main.tex
add acknowledgemtn
Add introduction

@@
 \begin{keywords}{Keywords:}
-Provide up to six keywords, separated by commas.
+Internet of Things (IoT), reinforcement learning.
 \end{keywords}
@@
 
-\section{About CAIAC}
+\section{Introduction}
 
-The Canadian Artificial Intelligence Association / Association Pour L'Intelligence Artificielle au Canada started out of a workshop in the Spring of 1973 and was originally named as the Canadian Society for Computational Studies of Intelligence/Societé canadienne pour études d'intelligence par ordinateur (CSCSI/SCEIO) and is probably the oldest organized association of its kind.
-The first formal CSCSI/SCEIO conference happened in 1976.
-The conference was held biannually for several years until it became an yearly event in 2000.
-In 2008, CSCSI/SCEIO was officially renamed to CAIAC.
+There is an estimated 1.5 million beef cows in Alberta with a portion of those being free-range cattle who are allowed to graze on large pastures for extended periods of time.
+Many farmers and researchers are interested in tracking movement of these cattle to asess the health of the herd, and to monitor the location of the cattle.
+These trackers take the form of collars, or ear tags and are equipped with sensors, batteries, and communication modules.
+The installation of the trackers is often a time consuming process so a tracker is expected to operate for extended periods of time (months) without human intervention.
+Given the size of the devices and the environment that they are expected to operate in, there is a limited amount of power budget available to the device and it beocmes a balancing act to ensure contiunuous operation, transmission and data collection.
+Reasearchers would like a device that transmits information at high rates as higher fidelity information allows them to make better conclusions about the state of the herd.
 
-More about our history can be found at \url{https://www.caiac.ca/en/history-of-caiac}.
+The paradigm of Internet of Things (IoT) devices has emerged over the last two decade as a way to deploy compute and intelligence throughout the environment.
+% The number of IoT devices is expected to grow to 75 billion by 2025 \cite{gubbiInternetThingsIoT2013}.
+The devices are often small, low power, and have limited communication capabilities, but provide insights into the environment that they are deployed in.
+The cattle trackers are an example of an IoT device that is deployed in the environment to provide insights into the health and location of the herd.
+IoT devices combine sensors, communication modules, storage, and compute to provide insights into the environment that they are deployed in.
+IoT devices come as networks or individual devices and in the case of a cattle tracker, ideally a tracker would be self-sufficient as a stand-alone item.
+With the spread of the internet and Starlink-like services, it is possible to have a tracker that is connected to the internet and can send data to the cloud, which enables the farmer to monitor the herd from anywhere in the world.
+Again we have the challenge of limited power budget and the need to transmit data at high rates.
+The power budget can be extended with the use of solar power. 
+Solar power is susceptible to weather conditions and the amount of sunlight that is available to the device, which in the case of cattle trackers also depends on the location/orientation of the cattle.
 
-\section{Submission Details}
-\label{submission}
-We invite submissions of both long and short papers.
-Long papers must be no longer than \textbf{12 pages}, and Short papers must be no longer than \textbf{6 pages}, including references, formatted using the conference template.
+Reinforcement learning is a branch of machine learning that is allows an agent to learn optimum actions to take in an environment to maximize a reward.
+The agent learns these actions by interacting with the environment and receiving feedback on the actions that it takes.
+Reinforcement learning has been used in a variety of applications such as playing games, optimizing energy consumption, and optimizing the operation of IoT devices.
+Machine learning deployed on IoT devices is somtimes referred to as TinyML or AIoT [Check this, since this is Copilot generated text].
+On one hand the addition of intelligence to the device allows it to make better decisions, but on the other hand the addition of intelligence to the device increases the power consumption of the device.
+Q-learning is a reinforcement learning algorithm that is used to optimize the actions that an agent takes in an environment and has been used in a variety of applications such as optimizing the operation of IoT devices.
 
-Papers submitted to the conference must contain original work that has not already been published, accepted for publication, or under review by a journal or another conference.
-Submissions will go through a \textbf{double-blind} review process.
-Authors are responsible for making sure submissions are anonymized.
+In this paper we propose a reinforcement learning enabled cattle tracker prototype that is able to optimize the energy consumption of the cattle tracker.
+The cattle tracker alternates between sleeping to conserve power and decision making. 
+It is faced with three decision points:
 
-A ``Best Paper Award'' and a ``Best Student Paper Award'' will be given at the conference respectively to the authors of each best paper, as judged by the Best Paper Award Selection Committee.
+\begin{itemize}
+  \item when to transmit data (high power consumption)
+  \item when to collect and store data (medium power consumption)
+  \item when to do nothing instead (lowest power consumption)
+\end{itemize}
 
-\section{Publication and Registration}
-\label{pub}
-Since 2020, the conference proceedings are published through PubPub open access\footnote{\url{https://www.pubpub.org/}} and indexed DBLP, ACM, Google Scholar, and other services that access Pubpub publications.
+Data transmission is needed to check in and ensure that the device and the cattle is still operational, but consumes a large amount of power.
+Data collection and storage is needed to ensure uninterrupted data collection, and consumes a medium amount of power.
+Doing nothing consumes the least amount of power, but the device incurs gaps in the data collection, which should only be done to avoid draining the battery to an empty state.
+Doing nothing will get the device through the night/day to ensure that it doesn't end up in a weird state by discharging the battery completely.
 
-At least one author of each accepted paper is required to attend the conference to present the work. The authors must agree to this requirement prior to submitting their paper for review.
+\section{Related Work}
 
-In addition, the corresponding author of each paper, acting on behalf of all of the authors of that paper, must complete and sign a Consent-to-Publish form.
-The corresponding author signing the copyright form should match the corresponding author marked on the paper.
+IoT Sensors are organized as follows:
+\begin{itemize}
 
-\section{Figures and Tables}
+\item Edge devices are the sensors that collect data.
+\item Fog layer which include edge computing devices and IoT gateways.
+\item Cloud layer which includes the cloud servers.
+\end{itemize}
 
-This simple \LaTeX\xspace template has been tested with \texttt{latex} and with \texttt{pdflatex} commands, directly from the command line or through \texttt{latexmk}.
-We ask that you place captions below tables and figures as in the examples below.
+The challenges of IoT sensors are \cite{chenDeepReinforcementLearning2021}:
 
-Pubpub uses a fairly complete \LaTeX\xspace engine that is likely to handle your source files without problems.
-Please check \url{https://help.pubpub.org/pub/latex-compatibility} if you require specialized packages or non-standard customization.
+- limited battery life
 
-\begin{figure}[ht]
-  \centering
-  \ifpdf
-    \includegraphics[scale=0.6]{figs/sample_fig.png}
-  \else
-    \includegraphics[scale=0.6,natwidth=330,natheight=120]{figs/sample_fig.png}
-  \fi
-  \caption{Captions under figures, please.}
-  \label{fig:sample_fig}
-\end{figure}
+- limited processing power
 
-\begin{table}[h]
-\begin{tabular}{|l|l|}
-\textbf{Header one} & \textbf{Header two} \\ 
-\hline
- Some text & More text
-\end{tabular}
-\vspace{0.2cm}
-\caption{Captions under tables, please.}
-\label{tab:important}
-\end{table}
+- limited storage
+
+IoT Networks have been used in a variety of applications
+
+- In \cite{hribarUsingDeepQLearning2019} the authors use deep Q-learning to extend the life of generic sensors networks with non-rechargable batteries by monitoring and optimizing a network of sensors.
+
+- In \cite{yamsaniIoTBasedLivestockMonitoring2024} the authors use IoT networks to monitor the health of livestock across the physical, network, edge and cloud layers.
+
+Contrary to the above works, we propose a single sensor agent, that decides when to transmit, collect and store, or send data to the cloud.
+This is done by using reinforcement learning to optimize the energy consumption of the sensor.
+
+\section{System Design and Architecture}
+
+\section{Implementation}
+
+\section{Results}
 
@@
 \section*{Acknowledgements}
-This section can be left blank during double-blind review.
+We would like to acknowledege the funding provided by NeurAlbertaTech for this project.
 
@@
 %\label{appendix-customize-this-label}
-Binomial theorem: \cite{abramowitz1948handbook}
+Binomial theorem: \cite{hribarUsingDeepQLearning2019}
 \begin{equation}

### 2025-02-12 paper/main.tex
Update main.tex

@@
 
-\section{Using Overleaf}
+The developed reinforcement learning-enabled cattle tracker prototype was successfully assembled and tested.
+The prototype consists of key components such as a solar panel, battery, communication module, and processing unit, as shown in Figure~\ref{fig:prototype}.
 
-If you are reading this you have already downloaded the zip file with the template (\textbf{CANAI \conferenceyear\xspace LaTex Template.zip}). All you need to do is to upload it into Overleaf:
+\begin{figure}[htbp]
+  \centering
+  \includegraphics[width=0.8\textwidth]{./figs/prototype.png}
+  \caption{Prototype of the reinforcement learning-enabled cattle tracker.
+The device consists of a solar panel for power harvesting, a rechargeable battery, and an embedded microcontroller for decision-making and communication.}
+  \label{fig:prototype}
+\end{figure}
 
-\begin{enumerate}
-\item Go to \url{overleaf.com}.
-\item Click on new project (top left menu).
-\item Choose the \textbf{upload project} option.
-\item Upload the \textbf{zip file} to Overleaf.
-\end{enumerate}
+The system was evaluated based on its power consumption, data transmission efficiency, and overall performance in real-world conditions.
+The reinforcement learning algorithm optimized the power usage by balancing data collection, storage, and transmission to ensure continuous operation.
+
+Additionally, the power levels of the device over time were analyzed to assess energy efficiency.
+Figure~\ref{fig:power_levels} shows the variation in power consumption during different operational phases.
+
+\begin{figure}[htbp]
+    \centering
+    \includegraphics[width=0.8\textwidth]{./figs/power_levels.png}
+    \caption{Power consumption levels of the cattle tracker over time. Different colors represent different operational phases such as data transmission, storage, and idle states.}
+    \label{fig:power_levels}
+\end{figure}
+
+\section{Conclusion}
 

### 2025-02-13 paper/main.tex
update
Add some related work and start the System Design and Architecture section

@@
 Reasearchers would like a device that transmits information at high rates as higher fidelity information allows them to make better conclusions about the state of the herd.
+Sensors data that researchers are interested in includes GPS coordinates, accelerometer data snapshots, and temperature data.
 
@@
 Reinforcement learning has been used in a variety of applications such as playing games, optimizing energy consumption, and optimizing the operation of IoT devices.
-Machine learning deployed on IoT devices is somtimes referred to as TinyML or AIoT [Check this, since this is Copilot generated text].
+Machine learning deployed on IoT devices is somtimes referred to as Artificial intelligence of Things (AIoT)\cite{yamsaniIoTBasedLivestockMonitoring2024}, which includes a broad range of applications.
+The TinyML paradigm is a subset of AIoT and focuses on machine learning on devices with highly constrained resources, which often limits it to inference only\cite{rayReviewTinyMLStateoftheart2022}.
 On one hand the addition of intelligence to the device allows it to make better decisions, but on the other hand the addition of intelligence to the device increases the power consumption of the device.
@@
 
-In this paper we propose a reinforcement learning enabled cattle tracker prototype that is able to optimize the energy consumption of the cattle tracker.
+In this paper we present the development of a prototype for a cattle tracker that uses q-learning to optimize the power consumption of the device.
+The prototype is a stand-alone device that is powered by a solar panel and is able to make decisions on when to transmit data to the cloud, when to collect and store data, and when to do nothing.
 The cattle tracker alternates between sleeping to conserve power and decision making. 
@@
 \section{Related Work}
+% IoT and Cloud
+The main challenges of IoT sensors are \cite{chenDeepReinforcementLearning2021} are (i) limited battery life, (ii) limited processing power, and (iii) limited storage.
+IoT devices are often limited to inference only, as training a model on the device is computationally expensive.
+Luckily, the cloud can be used to train the model and the model can be deployed on the device.
 
-IoT Sensors are organized as follows:
-\begin{itemize}
+%Cattle
+Cattle tracking with IoT devices is a growing field of research. In \cite{yamsaniIoTBasedLivestockMonitoring2024} the authors propose a network of IoT devices to monitor cattle.
+The network consists of edge devices that are attached to the cattle, fog devices that are attached to the barn, and cloud devices that are attached to the cloud. 
+In \cite{duttaMOOnitorIoTBased2022} the authors present an IoT device that clasifies the behaviour of the cattle based on data from accelerometer, temperature via XGBoost and Random Forest classifiers.
+Their device lasts for an estimated 10 hours on a single charge.
 
-\item Edge devices are the sensors that collect data.
-\item Fog layer which include edge computing devices and IoT gateways.
-\item Cloud layer which includes the cloud servers.
-\end{itemize}
-
-The challenges of IoT sensors are \cite{chenDeepReinforcementLearning2021}:
-
-- limited battery life
-
-- limited processing power
-
-- limited storage
-
-IoT Networks have been used in a variety of applications
-
-- In \cite{hribarUsingDeepQLearning2019} the authors use deep Q-learning to extend the life of generic sensors networks with non-rechargable batteries by monitoring and optimizing a network of sensors.
-
-- In \cite{yamsaniIoTBasedLivestockMonitoring2024} the authors use IoT networks to monitor the health of livestock across the physical, network, edge and cloud layers.
-
-Contrary to the above works, we propose a single sensor agent, that decides when to transmit, collect and store, or send data to the cloud.
-This is done by using reinforcement learning to optimize the energy consumption of the sensor.
+% Reinforcement Learning
+In \cite{hribarUsingDeepQLearning2019} the authors use deep Q-learning to extend the life of generic sensors networks with non-rechargable batteries. 
+The authors take advantage of corelation between sensor measurements to determine the rate at which sensors should transmit to maximize battery life.
 
 \section{System Design and Architecture}
+The design of a reinforcement learning enabled cattle tracker requires an iterative design process that includes assumptions and validations.
+Due to the iterateive nature, good pipelining processes are required to ensure that the design is validated at each step.
+First, a model was created with assumed power consumption and generation numbers to validate the agent.
+Next the prototype was constructed and the power consumption measured to update the model.
+The prototype was also used to collect and measure days for power generation under stationary conditions.
+The agent was then updated with the new power consumption and trained with randomly sampled data from the model.
+The agent was then deployed on the prototype and the power consumption was measured to validate the agent.
+
+\subsection{Q-learning}
+
+In Q-learning an agent learns over many iteration how to choose an action $a$, given a state $s$, to maximize the reward $r$.
+
+\begin{equation}
+  Q(s,a) \leftarrow Q(s,a) + \alpha \left( r + \gamma \max_{a'} Q(s',a') - Q(s,a) \right)
+\end{equation}
+
+\subsection{Hardware}
+The hardware prototype is shown in \ref{fig:prototype} and consists of a solar panel, battery, communication module, and processing unit.
+The specifications are shown in table \ref{tab:hardware_inventory}.
+
+\begin{table}[h!]
+  \centering
+  \caption{Hardware Inventory}
+  \begin{tabular}{|l|l|l|l|}
+  \hline
+  \textbf{Hardware} & \textbf{Serial Number} & \textbf{Purpose} & \textbf{Specs} \\ \hline
+  Arduino MKR WiFi & SN12345 & Connectivity & WiFi-enabled microcontroller board \\ \hline
+  Arduino MKR GPS  & SN67890 & Location Tracking & Integrated GPS module \\ \hline
+  DFRobot Solar Power Manager & SN11223 & Power Regulation & Manages and distributes solar cell and battery \\ \hline
+  Solar Panel 70x100 mm & SN44556 & Energy Generation & 70x100 mm panel, 5W output \\ \hline
+  Sparkfun Fuel Gage & SN77889 & Fuel Level Monitoring & Analog sensor for fuel measurement \\ \hline
+  LiPo Battery & HXJNLDC & Energy Storage & 3.7V, 2000mAh capacity \\ \hline
+  \end{tabular}
+  \label{tab:hardware_inventory}
+  \end{table}
+
+\subsection{Firmware}
+
+\subsection{Cloud}
 
@@
   \centering
-  \includegraphics[width=0.8\textwidth]{./figs/prototype.png}
+  \includegraphics[width=0.5\textwidth]{./figs/prototype_clean.png}
   \caption{Prototype of the reinforcement learning-enabled cattle tracker.

### 2025-02-13 paper/main.tex
Add more implementation

@@
 
-In this paper we present the development of a prototype for a cattle tracker that uses q-learning to optimize the power consumption of the device.
-The prototype is a stand-alone device that is powered by a solar panel and is able to make decisions on when to transmit data to the cloud, when to collect and store data, and when to do nothing.
+In this paper we present the development of a self-sufficient cattle tracker prototype capable of monitoring a single cow for extended periods of time.
+The tracker should record and collect data at regular intervals and transmit the data to the cloud with a maximum specified delay.
+Solar power can be utilized to extend the power budget of the device.
+Given that the device is capable of generating power, it should use excess power to transmit data more frequently, but the exact strategy is not trivial to determine.
+Reinforcement learning is used to optimize the power consumption of the device by making decisions on when to transmit data, when to collect and store data, and when to do nothing.
+
 The cattle tracker alternates between sleeping to conserve power and decision making. 
@@
 \section{System Design and Architecture}
-The design of a reinforcement learning enabled cattle tracker requires an iterative design process that includes assumptions and validations.
-Due to the iterateive nature, good pipelining processes are required to ensure that the design is validated at each step.
+
+The design of an AIoT device is an iterative process, that loops between simulations, hardware development and measure.
 First, a model was created with assumed power consumption and generation numbers to validate the agent.
@@
 
-In Q-learning an agent learns over many iteration how to choose an action $a$, given a state $s$, to maximize the reward $r$.
+Central to q-learning is the action-value function $Q(s,a)$, which is the expected return of taking action $a$ in state $s$.
+The agent learns the optimal policy by updating the action-value function based on the reward that it receives over many training iterations
 
@@
 
+There are three main components to the q-learning algorithm: the state, the action, and the reward.
+
+The state for the agent is a combination of (i) the time of day (ii) the power level of the battery (iii) the number of messages stored in the device.
+The time of day is discretized into 30 minute intervals, which creates 48 possible values. 
+The power level is discretized into 20 levels. 
+The number of messages stored in the device is discretized into 5 levels.
+The action for the agent is one of three (i) transmit data (ii) collect and store data (iii) do nothing.
+
+\begin{equation}
+  \mathcal{A} = \left\{
+  \begin{array}{l}
+  a_1 = \text{``transmit data''}, \\
+  a_2 = \text{``collect and store data''}, \\
+  a_3 = \text{``do nothing''}
+  \end{array}
+  \right.
+  \end{equation}
+
+The action-value $Q(a,s)$ function can take many forms. 
+For example there is deep Q-learning where the action-value function is approximated by a neural network.
+Due to the resource constraints, the action-value function is approximated by a table in this case.
+A table is an attractive option for inference only applications, as it is fast to access $O(1)$, requires little computation and only requires a sufficiently large memory.
+An additional benefit is that the table can be easily updated from the cloud as individual entries can be updated to reflect new conditions.
+
 \subsection{Hardware}
-The hardware prototype is shown in \ref{fig:prototype} and consists of a solar panel, battery, communication module, and processing unit.
-The specifications are shown in table \ref{tab:hardware_inventory}.
+The hardware prototype is shown in \ref{fig:prototype} and consists of an Arduino microcontroller, a GPS device,  a solar array, a power distribution system, and a battery.
+The detailed specifications are shown in table \ref{tab:hardware_inventory}.
 
@@
 
+The prototype-level hardware experiences higher power consumption during sleeping, which results in shorter battery life. 
+A side-benefit of this larger power consumption is a faster feedback cycle for development and troubleshooting. 
+A custom-build PCB is planned for the production-level hardware to reduce power consumption during sleep and ensure robustness of the final device.
+
 \subsection{Firmware}
+The firmware for the device is written in C++ and is responsible for managing the power consumption of the device.
+The main loop runs at a 30 minute interval and the device alternates between sleeping and decision making.
+At each decision point the device reads the time, the power level of the battery, and the number of messages stored in the device and retrivees a recommended action.
+The device then takes the recommended action and updates the power level of the battery.
+
+The device is limited in its memory to store the policy.
+To pair down the table, the highest action for each state was selected.
+the policy was then stored as values that were bit-shifted to fit into a 32-bit integer.
 
 \subsection{Cloud}
+The device transmits data to Azure IoT Central.
+IoT Central is an attractive option as it allows the registering of the device for security purposes, the creation of dahsboards, and the creation of digital twins.
+The digital twin is a digital representation of the device and can be used to monitor the device and update the state of the device.
+The cloud enables two-way communication, which allows the device to receive updates and new policies from the cloud whenever it is connected.
+For example the cloud can update the device with a new policy for a new season, or a new policy for a new location that affects the solar gain.
 
 \section{Implementation}
+
+%TODO: Add sample power consumptions graphs
 

### 2025-02-13 paper/main.tex
Additional edits

@@
 Many farmers and researchers are interested in tracking movement of these cattle to asess the health of the herd, and to monitor the location of the cattle.
-These trackers take the form of collars, or ear tags and are equipped with sensors, batteries, and communication modules.
-The installation of the trackers is often a time consuming process so a tracker is expected to operate for extended periods of time (months) without human intervention.
-Given the size of the devices and the environment that they are expected to operate in, there is a limited amount of power budget available to the device and it beocmes a balancing act to ensure contiunuous operation, transmission and data collection.
+Reasearchers are interested in behaviour of the herd like feeding, rumination, and resting and past research has shown that this behaviour can be correlated to sensor data from accelerometer \cite{unoldIoTBasedCowHealth2020}.
 Reasearchers would like a device that transmits information at high rates as higher fidelity information allows them to make better conclusions about the state of the herd.
-Sensors data that researchers are interested in includes GPS coordinates, accelerometer data snapshots, and temperature data.
+These trackers take the form of collars, leg bands, or ear tags and are equipped with sensors, batteries, and communication modules.
+The installation of the trackers is a time consuming process so a tracker is expected to operate for extended periods of time (months) without human intervention.
+Given the size of the devices and the environment that they operate in, there is a limited power budget available to the device and it beocmes a balancing act to ensure contiunuous operation, transmission and data collection.
 
-The paradigm of Internet of Things (IoT) devices has emerged over the last two decade as a way to deploy compute and intelligence throughout the environment.
+The paradigm of Internet of Things (IoT) devices has emerged over the last two decades as a way to deploy sensors, communication modules, storage, compute, and intelligence throughout the environment.
 % The number of IoT devices is expected to grow to 75 billion by 2025 \cite{gubbiInternetThingsIoT2013}.
 The devices are often small, low power, and have limited communication capabilities, but provide insights into the environment that they are deployed in.
-The cattle trackers are an example of an IoT device that is deployed in the environment to provide insights into the health and location of the herd.
-IoT devices combine sensors, communication modules, storage, and compute to provide insights into the environment that they are deployed in.
-IoT devices come as networks or individual devices and in the case of a cattle tracker, ideally a tracker would be self-sufficient as a stand-alone item.
-With the spread of the internet and Starlink-like services, it is possible to have a tracker that is connected to the internet and can send data to the cloud, which enables the farmer to monitor the herd from anywhere in the world.
-Again we have the challenge of limited power budget and the need to transmit data at high rates.
-The power budget can be extended with the use of solar power. 
-Solar power is susceptible to weather conditions and the amount of sunlight that is available to the device, which in the case of cattle trackers also depends on the location/orientation of the cattle.
+Cattle trackers are an example of an IoT device that is deployed in the environment to provide insights into the health and location of the herd.
+These cattle trackers are strating to replace labour intensive manual observation with automated systems that can monitor the herd 24-7 \cite{unoldIoTBasedCowHealth2020} \cite{moutaouakilDigitalFarmingSurvey2023}.
+IoT devices can operate as part of a network or as standalone units. 
+For cattle tracking, a self-sufficient device that connects to the internet via cellular networks such as GSM, LTE, or even satellite services like Starlink is ideal, allowing farmers to monitor their herds remotely.
+However, these devices face power constraints due to the high data transmission rates required; solar panels can extend battery life, though their effectiveness depends on weather conditions and the cattle's orientation and location.
 
-Reinforcement learning is a branch of machine learning that is allows an agent to learn optimum actions to take in an environment to maximize a reward.
-The agent learns these actions by interacting with the environment and receiving feedback on the actions that it takes.
+Reinforcement learning is a branch of machine learning that allows an agent to learn optimum actions to take in an environment to maximize a reward.
+The agent learns these actions by interacting with the environment and receiving rewards for those actions.
 Reinforcement learning has been used in a variety of applications such as playing games, optimizing energy consumption, and optimizing the operation of IoT devices.
-Machine learning deployed on IoT devices is somtimes referred to as Artificial intelligence of Things (AIoT)\cite{yamsaniIoTBasedLivestockMonitoring2024}, which includes a broad range of applications.
+The combination of reinforcement learning and IoT devices is an attractive option as it allows the device to make decisions without the need for human intervention.
+THe combvination of AI and IoT is sometimes called the Artificial intelligence of Things (AIoT)\cite{yamsaniIoTBasedLivestockMonitoring2024}, which includes a broad range of applications.
 The TinyML paradigm is a subset of AIoT and focuses on machine learning on devices with highly constrained resources, which often limits it to inference only\cite{rayReviewTinyMLStateoftheart2022}.
@@
 
-In this paper we present the development of a self-sufficient cattle tracker prototype capable of monitoring a single cow for extended periods of time.
-The tracker should record and collect data at regular intervals and transmit the data to the cloud with a maximum specified delay.
-Solar power can be utilized to extend the power budget of the device.
-Given that the device is capable of generating power, it should use excess power to transmit data more frequently, but the exact strategy is not trivial to determine.
-Reinforcement learning is used to optimize the power consumption of the device by making decisions on when to transmit data, when to collect and store data, and when to do nothing.
+In this paper we present the development of a cattle tracker prototype capable of data collection for extended periods of time.
+The main controbution of this paper is the development of a reinforcement learning agent that optimizes the power consumption of the device by allowing the agent to make decisions on when to transmit data, when to collect and store data, and when to do nothing.
+The hardware prototype uses solar power to extend the power budget of the device and the agent is trained to take advantage of excess power availability to increase the data transmission rate.
 
@@
 The main challenges of IoT sensors are \cite{chenDeepReinforcementLearning2021} are (i) limited battery life, (ii) limited processing power, and (iii) limited storage.
-IoT devices are often limited to inference only, as training a model on the device is computationally expensive.
-Luckily, the cloud can be used to train the model and the model can be deployed on the device.
+Devices that fall under the TinyML paradigm are often limited to inference only, as training a model on the device is computationally expensive.
+The addition of cloud communication opens up the possibility of training the model in the cloud and deploying a new model to the device that adapts to changes in the environment.
 
 %Cattle
-Cattle tracking with IoT devices is a growing field of research. In \cite{yamsaniIoTBasedLivestockMonitoring2024} the authors propose a network of IoT devices to monitor cattle.
-The network consists of edge devices that are attached to the cattle, fog devices that are attached to the barn, and cloud devices that are attached to the cloud. 
+Agriculture and cattle tracking commonly employ IoT devices for monitoring. 
+In \cite{yamsaniIoTBasedLivestockMonitoring2024} the authors propose a network of IoT devices to monitor cattle, with edge devices for network communication and cloud services for data storage and analysis.
+Multiple standalone devices have also been developed for cattle tracking, with many focusing on the sensor output of the device.
 In \cite{duttaMOOnitorIoTBased2022} the authors present an IoT device that clasifies the behaviour of the cattle based on data from accelerometer, temperature via XGBoost and Random Forest classifiers.
 Their device lasts for an estimated 10 hours on a single charge.
+In \cite{unoldIoTBasedCowHealth2020} the authors present the CowMonitor an IoT device uses Bluetooth Low Energy (BLE) for communication with a hub and sends accelerometer and magnetometer data to the cloud.
+They transmit data without acknowledgement for oppurtunistic capture and lasts 5 years without battery replacement, but required the deployment of BLE hubs and WiFi ground stations thoughout the pasture.
 
 % Reinforcement Learning
+The combination of intelligence and IoT devices allows a powerful solution to add intelligence to the edge.
 In \cite{hribarUsingDeepQLearning2019} the authors use deep Q-learning to extend the life of generic sensors networks with non-rechargable batteries. 
@@
 An additional benefit is that the table can be easily updated from the cloud as individual entries can be updated to reflect new conditions.
+
+It is not trivial to determine the reward values for the agent to mximize data collection.
+Manual heuristics can be used to determine the reward values, but that is not efficient and may not be optimal.
+For the agent we consider the reward values to be hyperparameters that can be tuned to maximize the data collection.
+We use a Bayesian optimization framework—specifically, the Gaussian Process-based minimization function \verb|gp\_minimize| from the \verb|scikit-optimize| library to automatically search the hyperparameter space defined for each reward component (e.g., reward\_power\_loss, reward\_power\_multiplier, reward\_action\_0, reward\_action\_1, reward\_action\_2, and reward\_message\_count).
+By repeatedly running training sessions with slightly varied random seeds and averaging the performance metrics, our objective function (which returns the negative median sent messages) guides the search toward the parameter set that maximizes data collection
+This automated approach saves time compared to manual tuning efforts and has shown to be effective in optimizing the reward values.
 

### 2025-02-13 paper/main.tex
Reformatted ref and added Conclusion

@@
 \section{Conclusion}
+In this study, we developed a reinforcement learning-enabled cattle tracker prototype designed to optimize power consumption while ensuring efficient data transmission and collection.
+The device successfully leverages reinforcement learning to dynamically adjust its operation based on available energy, enhancing battery life and operational efficiency.
+By incorporating solar power and an intelligent decision-making framework, the prototype demonstrates a feasible approach to long-term, autonomous cattle tracking.
+Future work will focus on improving the efficiency of the reinforcement learning model by integrating real-time adaptation to environmental changes.
+Additional experiments in diverse field conditions will provide further validation of the system's robustness.
+Further refinements to the hardware, such as low-power components and custom PCB design, will enhance reliability and longevity.
+Moreover, expanding the cloud-based digital twin functionality can improve remote monitoring and predictive analytics for herd management.
 

### 2025-02-13 paper/main.tex
Conclusion & Abstract added. Ref reformatted.

@@
 \begin{abstract}
-This is just a sample \LaTeX template. As usual, most of the customization for the proceedings is done in the \texttt{cai.cls} file. 
+The increasing demand for precision livestock monitoring has led to the development of Internet of Things (IoT)-enabled solutions for real-time tracking and health assessment of cattle.
+In this study, we present the development of a reinforcement learning-enabled cattle tracker prototype designed to optimize power consumption while ensuring continuous data collection and transmission.
+The prototype integrates IoT sensors, solar power harvesting, and reinforcement learning-based decision-making to dynamically balance data transmission, storage, and energy conservation.
+The reinforcement learning agent autonomously determines the optimal transmission schedule based on power availability, maximizing data efficiency while prolonging battery life.
+Experimental evaluations demonstrate the prototype's capability to sustain long-term operation under real-world environmental constraints.
+The results highlight the feasibility of integrating machine learning with IoT devices for adaptive, energy-efficient cattle monitoring.
+Our findings contribute to the broader field of smart agriculture by demonstrating how reinforcement learning can be leveraged to enhance IoT-based livestock management.
+Future work will focus on optimizing the decision-making model for diverse environmental conditions and scaling the system for larger deployments.
+
 \end{abstract}
@@
 \begin{keywords}{Keywords:}
-Internet of Things (IoT), reinforcement learning.
+Internet of Things (IoT), Reinforcement Learning, Cattle Monitoring.
 \end{keywords}

### 2025-02-14 paper/main.tex
Acknowledgements can be left blank during double-blind review.
Update the q-learning section

@@
 
-Central to q-learning is the action-value function $Q(s,a)$, which is the expected return of taking action $a$ in state $s$.
-The agent learns the optimal policy by updating the action-value function based on the reward that it receives over many training iterations
+Central to q-learning is the action-value function $Q(s,a)$, which is the expected cumulative reward of taking action $a$ in state $s$.
+The agent learns the optimal policy by updating the action-value function based on the reward that it receives over many training iterations according to:
 
@@
 \end{equation}
+where $\alpha$ is the learning rate, $r$ is the reward, $\gamma$ is the discount factor, and $s'$ is the next state after taking action $a$ in state $s$.
 
-There are three main components to the q-learning algorithm: the state, the action, and the reward.
+The state of the agent is a combination of the time of day $t$, the power level of the battery $p$, and the number of messages queued on the device $m$.
+The time of day informs the agent of the likelyhood of power generation and was discretized to 48 bins.
+The maximum power state was discretized into 100 power levels and the maximum number of messages was 5, forcing the agent to check in every 2.5 hours or otherwise risk losing messages.
+This resulted in 24,800 possible states for the agent to be in.
 
-The state for the agent is a combination of (i) the time of day (ii) the power level of the battery (iii) the number of messages stored in the device.
-The time of day is discretized into 30 minute intervals, which creates 48 possible values. 
-The power level is discretized into 20 levels. 
-The number of messages stored in the device is discretized into 5 levels.
-The action for the agent is one of three (i) transmit data (ii) collect and store data (iii) do nothing.
+The agent can choose one of three actions (i) transmit data (ii) collect and store data (iii) do nothing.
 
@@
   \end{equation}
+Turning the action-value function into a $4800 \times 3$ table.
 
-The action-value $Q(a,s)$ function can take many forms. 
-For example there is deep Q-learning where the action-value function is approximated by a neural network.
-Due to the resource constraints, the action-value function is approximated by a table in this case.
-A table is an attractive option for inference only applications, as it is fast to access $O(1)$, requires little computation and only requires a sufficiently large memory.
+A tabular representation of the action-value function $Q(a,s)$ was used as it turns the model inference into a table lookup with $O(1)$ complexity.
+Alternatives to the tabular representation include deep Q-learning, which uses a neural network to approximate the action-value function, and is more computationally expensive.
+A tabular representation only requires a sufficiently large memory to store the table.
 An additional benefit is that the table can be easily updated from the cloud as individual entries can be updated to reflect new conditions.
 
-It is not trivial to determine the reward values for the agent to mximize data collection.
-Manual heuristics can be used to determine the reward values, but that is not efficient and may not be optimal.
-For the agent we consider the reward values to be hyperparameters that can be tuned to maximize the data collection.
-We use a Bayesian optimization framework—specifically, the Gaussian Process-based minimization function \verb|gp\_minimize| from the \verb|scikit-optimize| library to automatically search the hyperparameter space defined for each reward component (e.g., reward\_power\_loss, reward\_power\_multiplier, reward\_action\_0, reward\_action\_1, reward\_action\_2, and reward\_message\_count).
-By repeatedly running training sessions with slightly varied random seeds and averaging the performance metrics, our objective function (which returns the negative median sent messages) guides the search toward the parameter set that maximizes data collection
-This automated approach saves time compared to manual tuning efforts and has shown to be effective in optimizing the reward values.
+The goal of the agent was to maximize the number of messages sent to the cloud, while minimizing the power consumption of the device.
+It is not trivial to determine the reward values for the agent to maximize data collection.
+We consider the reward values to be hyperparameters that can be tuned to maximize the data collection and used a Bayesian optimization framework from the the python \verb|scikit-optimize| package to automatically tune the hyperparameters.
+The hyper parameters could be nagative and they were:
+\begin{itemize}
+  \item reward\_power\_loss: the reward for losing power
+  \item reward\_power\_multiplier: the reward for having excess power at every time step
+  \item reward\_action\_0: the reward for transmitting data
+  \item reward\_action\_1: the reward for collecting and storing data
+  \item reward\_action\_2: the reward for doing nothing
+  \item reward\_message\_count: the reward for the number of messages sent when transmission occurred
+\end{itemize}
+
+We ran repeated training sessions over solar gain data collected from the physical prototype that guided us toward the parameter set that maximizes data collection.
+This automated approach was less labour intensive than manual tuning and has shown to be effective in optimizing the reward values.
 
@@
   Sparkfun Fuel Gage & SN77889 & Fuel Level Monitoring & Analog sensor for fuel measurement \\ \hline
-  LiPo Battery & HXJNLDC & Energy Storage & 3.7V, 2000mAh capacity \\ \hline
+  LiPo Battery & HXJNLDC & Energy Storage & 3.7V, 2100mAh capacity \\ \hline
   \end{tabular}
@@
 \section*{Acknowledgements}
-We would like to acknowledege the funding provided by NeurAlbertaTech for this project.
-
-%Appendixes go here
-\appendix
-
-\section{Example of math equation }
-%\label{appendix-customize-this-label}
-Binomial theorem: \cite{hribarUsingDeepQLearning2019}
-\begin{equation}
-(x+y)^n=\sum_{\substack{k=0}}^{n}\dbinom{n}{k}x^{n-k}y^k
-\end{equation}
 

### 2025-02-14 paper/main.tex
Updated Table 1 - Serial Num ad Specs

@@
   \textbf{Hardware} & \textbf{Serial Number} & \textbf{Purpose} & \textbf{Specs} \\ \hline
-  Arduino MKR WiFi & SN12345 & Connectivity & WiFi-enabled microcontroller board \\ \hline
-  Arduino MKR GPS  & SN67890 & Location Tracking & Integrated GPS module \\ \hline
-  DFRobot Solar Power Manager & SN11223 & Power Regulation & Manages and distributes solar cell and battery \\ \hline
-  Solar Panel 70x100 mm & SN44556 & Energy Generation & 70x100 mm panel, 5W output \\ \hline
-  Sparkfun Fuel Gage & SN77889 & Fuel Level Monitoring & Analog sensor for fuel measurement \\ \hline
-  LiPo Battery & HXJNLDC & Energy Storage & 3.7V, 2100mAh capacity \\ \hline
+  Arduino MKR WiFi 1010 & ABX00023 & Connectivity & WiFi-enabled microcontroller board \\ \hline
+  Arduino MKR GPS Shield & ASX00017 & Location Tracking & Integrated GPS module \\ \hline
+  DFRobot Solar Power Manager & TPX00046 & Power Regulation & Manages and distributes solar cell and battery \\ \hline
+  Solar Panel & TPX00181 & Energy Generation & 75x100x2.9 mm panel, 5W output \\ \hline
+  Sparkfun Fuel Gage & 1568-1273-ND & Fuel Level Monitoring & Analog sensor for fuel measurement \\ \hline
+  LiPo Battery & 2528 & Energy Storage & 3.7V, 2000mAh capacity \\ \hline
   \end{tabular}

### 2025-02-14 paper/main.tex
updated paper and a sample power consumption (with the GPS included)

--- 2025-02-14 paper/main.tex
@@
 
-The goal of the agent was to maximize the number of messages sent to the cloud, while minimizing the power consumption of the device.
+The goal of the agent is to maximize the number of messages sent to the cloud, while minimizing the power consumption of the device.
 It is not trivial to determine the reward values for the agent to maximize data collection.
@@
 
-We ran repeated training sessions over solar gain data collected from the physical prototype that guided us toward the parameter set that maximizes data collection.
+The agent was trained on a simulated environment with the power consumption and generation numbers from the prototype.
+At every step, the environment would update the power level of the battery, the time of day, and the number of messages queued on the device depending on the action taken by the agent.
+The actual measured battery capacity was 1300 mAh and every 30 minutes the device would consume 5.0 mAh in sleeping.
+A power transmission would consume an additional 1.6 mAh and a data collection would consume an additional 0.1 mAh.
+
+Sample solar gain data from 17 days in June in Edmonton was turned into an average sunny day and an average cloudy day. 
+The environment would switch between the two days to simulate the weather conditions with an 80\% chance of a cloudy day.
+These conditions ensured that there was not sufficient power available to transmit data at every time step and forced the agent to learn strategies to increase message count.
+
+The training run was repeated 4 times each with a different random seed at every iteration of the hyper parameter search.
+Once an optimum set of hyperparameters was found, the action-value function was deployed into an h-file that was used in the firmware of the device.
 This automated approach was less labour intensive than manual tuning and has shown to be effective in optimizing the reward values.
@@
   \centering
-  \caption{Hardware Inventory}
-  \begin{tabular}{|l|l|l|l|}
+  \caption{Hardware List for the Cattle Tracker Prototype}
+  \begin{tabular}{|l|l|l|}
   \hline
-  \textbf{Hardware} & \textbf{Serial Number} & \textbf{Purpose} & \textbf{Specs} \\ \hline
-  Arduino MKR WiFi & SN12345 & Connectivity & WiFi-enabled microcontroller board \\ \hline
-  Arduino MKR GPS  & SN67890 & Location Tracking & Integrated GPS module \\ \hline
-  DFRobot Solar Power Manager & SN11223 & Power Regulation & Manages and distributes solar cell and battery \\ \hline
-  Solar Panel 70x100 mm & SN44556 & Energy Generation & 70x100 mm panel, 5W output \\ \hline
-  Sparkfun Fuel Gage & SN77889 & Fuel Level Monitoring & Analog sensor for fuel measurement \\ \hline
-  LiPo Battery & HXJNLDC & Energy Storage & 3.7V, 2100mAh capacity \\ \hline
+  \textbf{Hardware} & \textbf{Serial Number} &  \textbf{Description} \\ \hline
+  Arduino MKR WiFi 1010 & ABX00023 &  WiFi-enabled microcontroller board \\ \hline
+  Arduino MKR GPS Shield & ASX00017 &  Integrated GPS module \\ \hline
+  DFRobot Solar Power Manager & TPX00046 &  Manages and distributes solar cell and battery \\ \hline
+  Solar Panel & TPX00181 &  75x100x2.9 mm panel, 1W output \\ \hline
+  Sparkfun Fuel Gage & 1568-1273-ND & Analog sensor for power measurement \\ \hline
+  LiPo Battery & 2528 &  3.7V, 2000mAh capacity \\ \hline
   \end{tabular}
@@
 
-The prototype-level hardware experiences higher power consumption during sleeping, which results in shorter battery life. 
-A side-benefit of this larger power consumption is a faster feedback cycle for development and troubleshooting. 
-A custom-build PCB is planned for the production-level hardware to reduce power consumption during sleep and ensure robustness of the final device.
+Testing on prototype-level hardware showed a high power consumption of 9.9 mA during sleeping due to the presence of multiple LEDs and integrated components that could not be disabled. 
+A custom-build PCB with isolation of every component should experience a sleep current on the order of $\mu A$ and is planned for the the next iteration hardware to reduce power consumption during sleep.
+An unexpected side-benefit of a prototype with large power consumption is a fast development feedback cycle where the prototype drains the battery in days, rather than weeks.
 
 \subsection{Firmware}
-The firmware for the device is written in C++ and is responsible for managing the power consumption of the device.
+The firmware for the protoype is written in C++ and is responsible for managing the power consumption of the device.
 The main loop runs at a 30 minute interval and the device alternates between sleeping and decision making.
@@
 The device is limited in its memory to store the policy.
-To pair down the table, the highest action for each state was selected.
-the policy was then stored as values that were bit-shifted to fit into a 32-bit integer.
+To pair down the table, only the highest action for each state was selected.
+The policy was then stored as values that were bit-shifted to fit into a 32-bit integer.
 
@@
 
-\section{Implementation}
-
-%TODO: Add sample power consumptions graphs
-
 \section{Results}
@@
   \centering
-  \includegraphics[width=0.5\textwidth]{./figs/prototype_clean.png}
+  \includegraphics[width=0.5\textwidth]{./figs/prototype.png}
   \caption{Prototype of the reinforcement learning-enabled cattle tracker.

### 2025-02-14 RL_nz/environment/custom_env.py
clean up environment code

@@ -2,13 +2,6 @@ import numpy as np
 
 from environment.device import Device
 
-def solar_intensity():
-    # Example implementation
-    # Replace with the actual calculation for solar intensity
-    intensity = np.zeros(48)
-    intensity[16:28] = 1
-    return intensity
-
 def get_solar_intensity(time_step: int, current_day: int = 0):
     cloudy_day = [0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0.34, 0, 0, 0, 0.34, 0, 0.34,0.34, 5.19, 5.197, 2.310, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0]
     sunny_day = [0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 13.2, 23.8, 49.6, 68.6, 58.0, 75.81, 74.89, 89.46, 86.70, 105, 105, 105, 105, 105, 97.3, 58.7, 58.7, 25.8, 20.3, 14.9, 20.4, 3.95, 9.43, 9.4, 3.9,0, 0, 0, 0, 0, 0, 0, 0, 0]
@@ -27,11 +20,9 @@ class CustomEnv:
         self.MAX_MESSAGES = config['max_messages'] 
         self._device = Device(power_max=config['max_power'],
                               rounding_factor=(100/self.N_POWER_LEVELS))
-                            #   rounding_factor=int(100/self.N_POWER_LEVELS))
         self._time = 0
         self.current_day = 0
-        self.cloudy_chance = 0.7
-        self.solar_intensity_cache = solar_intensity()*30
+        self.cloudy_chance = config.get('cloudy_chance', 0.8)
         
     def init_state(self, power_level: float=50, time: int=0, message_count: int=0):
         self._device.set_power_percentage(power_level)
@@ -42,7 +33,6 @@ class CustomEnv:
         return self._device.calculate_rounded_power_level()
         
     def encode_state(self):
-        # discretized_power_level = self.get_power()
         return self._encode_state(self.get_power(), self._time, self._device._message_queue_length)
 
     def _encode_state(self, power_level, time, message_count):

### 2025-02-14 RL_nz/solar_connection_search.py

@@ -45,7 +45,6 @@ def main():
             agent = DQNAgent(config['agent'], env)
 
             # Start training
-            env.cloudy_chance = 0.8
             message_count_list, power_list = train(env, agent, config['train'], plot_result_flag=False)
             message_counts.append(np.median(message_count_list[-1000:]))
             power_averages.append(np.median(power_list[-1000:]))

### 2025-02-14 paper/main.tex
Add additional information

@@
 The reinforcement learning agent autonomously determines the optimal transmission schedule based on power availability, maximizing data efficiency while prolonging battery life. 
+The agent is trained on measured prototype data and the reward function is tuned from a Bayesian optimization to maximize the message collection.
 Experimental evaluations demonstrate the prototype's capability to sustain long-term operation under real-world environmental constraints. 
@@
 
-The design of an AIoT device is an iterative process, that loops between simulations, hardware development and measure.
+The design of an AIoT device is an iterative process, that loops between simulations, hardware development, measurement and redeployment.
+It is critical to have an automated process in place that can be updated without manual intervention.
 First, a model was created with assumed power consumption and generation numbers to validate the agent.
@@
   \begin{array}{l}
-  a_1 = \text{``transmit data''}, \\
-  a_2 = \text{``collect and store data''}, \\
-  a_3 = \text{``do nothing''}
+    a_0 = \text{``do nothing''}
+    a_1 = \text{``collect and store data''}, \\
+    a_2 = \text{``collect and transmit data''}, \\
   \end{array}
@@
   \item reward\_power\_multiplier: the reward for having excess power at every time step
-  \item reward\_action\_0: the reward for transmitting data
+  \item reward\_action\_0: the reward for doing nothing
   \item reward\_action\_1: the reward for collecting and storing data
-  \item reward\_action\_2: the reward for doing nothing
+  \item reward\_action\_2: the reward for collecting and transmitting data
   \item reward\_message\_count: the reward for the number of messages sent when transmission occurred
@@
 \subsection{Firmware}
-The firmware for the protoype is written in C++ and is responsible for managing the power consumption of the device.
+The firmware for the prototype is written in C++ and is responsible for managing the power consumption of the device.
 The main loop runs at a 30 minute interval and the device alternates between sleeping and decision making.
@@
 \section{Results}
-
 The developed reinforcement learning-enabled cattle tracker prototype was successfully assembled and tested. 
 The prototype consists of key components such as a solar panel, battery, communication module, and processing unit, as shown in Figure~\ref{fig:prototype}.
+
+\subsection{Training Results}
+A baseline device that alternates between collection and transmission is used to compare the performance of the reinforcement learning agent.
+The results are shown in Table~\ref{tab:agent_comparison}.
+
+\begin{table}[h]
+  \centering
+  \begin{tabular}{lcc}
+      \toprule
+      & Baseline Agent & RL Agent \\
+      \midrule
+      Transmission Actions & 480 & 308 \\
+      Collection Actions & 480 & 652 \\
+      Sleep Actions & 0 & 0 \\
+      Avg. Message Count & 756.5 & 789.5 \\
+      Avg. Power Level & 34 & 76 \\
+      \bottomrule
+  \end{tabular}
+  \caption{Comparison of Baseline Agent and RL Agent}
+  \label{tab:agent_comparison}
+\end{table}
+
+The reinforcement learning agent outperforms the baseline agent by increasing the number of messages sent to the cloud by 4.3\% while maintaining a higher average power level by avoiding transmissions.
+
+\subsection{Deploymnet Results}
 

### 2025-02-14 paper/main.tex
Daily update

@@
 In this paper we present the development of a cattle tracker prototype capable of data collection for extended periods of time.
-The main controbution of this paper is the development of a reinforcement learning agent that optimizes the power consumption of the device by allowing the agent to make decisions on when to transmit data, when to collect and store data, and when to do nothing.
+It is not trivial to manually determine the optimal data collection strategy for the device under dynamic and complex conditions \cite{suttonReinforcementLearningIntroduction2020} so we propose the addition of reinforcement learning to automate the decision making process.
+The main contribution of this paper is the development of a reinforcement learning agent that optimizes the power consumption of the device by allowing the agent to make decisions on when to transmit data, when to collect and store data, and when to do nothing.
 The hardware prototype uses solar power to extend the power budget of the device and the agent is trained to take advantage of excess power availability to increase the data transmission rate.
+The IoT device likely fits other use cases, but the focus in this paper is on the cattle tracking application due to the requirements of collecting uninterrupted data, occasionally needing to check in, needing to operate for long periods of time, and the potential for a mobile solar panel, which results in drastically different power generation.
 
@@
 It is critical to have an automated process in place that can be updated without manual intervention.
-First, a model was created with assumed power consumption and generation numbers to validate the agent.
-Next the prototype was constructed and the power consumption measured to update the model.
-The prototype was also used to collect and measure days for power generation under stationary conditions.
-The agent was then updated with the new power consumption and trained with randomly sampled data from the model.
-The agent was then deployed on the prototype and the power consumption was measured to validate the agent.
+The use of reinforcement learning greatly simplifies the process and allows the device to adapt to new conditions without a manual update of collection strategies.
 
@@
 
+\begin{figure}[htbp]
+  \centering
+  \includegraphics[width=0.5\textwidth]{./figs/prototype.png}
+  \caption{Prototype of the reinforcement learning-enabled cattle tracker.
+The device consists of a solar panel for power harvesting, a rechargeable battery, and an embedded microcontroller for decision-making and communication.}
+  \label{fig:prototype}
+\end{figure}
+
 \subsection{Firmware}
@@
 
-The reinforcement learning agent outperforms the baseline agent by increasing the number of messages sent to the cloud by 4.3\% while maintaining a higher average power level by avoiding transmissions.
+The reinforcement learning agent outperforms the baseline agent by increasing the number of messages sent to the cloud by 4.3\% while maintaining a higher average power level by avoiding excess transmission actions.
 
 \subsection{Deploymnet Results}
-
-\begin{figure}[htbp]
-  \centering
-  \includegraphics[width=0.5\textwidth]{./figs/prototype.png}
-  \caption{Prototype of the reinforcement learning-enabled cattle tracker.
-The device consists of a solar panel for power harvesting, a rechargeable battery, and an embedded microcontroller for decision-making and communication.}
-  \label{fig:prototype}
-\end{figure}
 

### 2025-02-14 RL_nz/train.py

@@ -83,5 +83,6 @@ def train(env: CustomEnv, agent, train_config, plot_result_flag:bool=True, resul
         save_results(f'results/{result_prefix}reward_list.csv', reward_list)
         save_results(f'results/{result_prefix}legitimate_messages_list.csv', legitimate_messages_list)
         save_results(f'results/{result_prefix}power_list.csv', power_list)
+        save_results(f'results/{result_prefix}power_list.csv', action_list)
         
     return legitimate_messages_list, power_list

### 2025-02-14 paper/main.tex
Add graphs for prototype results

@@
 \documentclass[10pt]{cai}
-
+\captionsetup{font=small}
 \begin{document}
@@
 The reinforcement learning agent outperforms the baseline agent by increasing the number of messages sent to the cloud by 4.3\% while maintaining a higher average power level by avoiding excess transmission actions.
+The trained agent performed on average 2.12 collection actions for every transmission action, compared to a 1:1 ratio for the baseline agent.
 
-\subsection{Deploymnet Results}
+\subsection{Deployment Results}
+
+The trained agent was deployed and evaluated on the prototype hardware in an indoor environment over a 9 day period with occasional manual charging to ensure continuous operation.
+This test evaluated the long-term performance of the agent and hardware system and the results are shown in Figure~\ref{fig:prototype_result}.
+The hardware did experience
+
+\begin{figure}[h]
+  \centering
+  \begin{subfigure}{0.48\textwidth}
+      \centering
+      \includegraphics[width=\linewidth]{figs/prototype_power_level.png}
+      \caption{Prototype power level over a 9 day collection window}
+      \label{fig:prototype_power}
+  \end{subfigure}
+  \hfill
+  \begin{subfigure}{0.48\textwidth}
+      \centering
+      \includegraphics[width=\linewidth]{figs/battery_level_vs_collection_ratio.png}
+      \caption{Learned agent strategies as a function of power levels}
+      \label{fig:battery_vs_collection}
+  \end{subfigure}
+  \caption{Collected data from the prototype hardware highlighting a nonlinear power level discharge and the learned agent strategies.}
+  \label{fig:prototype_result}
+\end{figure}
+
+Figure~\ref{fig:prototype_power} highlights the nonlinear discharge curve of the battery over the 9 day period. 
+The discharge rate during sleeping is relatively constant and the nonlinearity is a result of the mapping function of voltage to remaining capacity.
+This nonlinearity poses a challenge to the agent, since it increases the error in future reward and remaining capacity estimation at lower power levels.
 
@@
 Figure~\ref{fig:power_levels} shows the variation in power consumption during different operational phases.
-
-\begin{figure}[htbp]
-    \centering
-    \includegraphics[width=0.8\textwidth]{./figs/power_levels.png}
-    \caption{Power consumption levels of the cattle tracker over time. Different colors represent different operational phases such as data transmission, storage, and idle states.}
-    \label{fig:power_levels}
-\end{figure}
 

### 2025-02-14 paper/main.tex
First draft will all sections filled in

@@
 
-The trained agent was deployed and evaluated on the prototype hardware in an indoor environment over a 9 day period with occasional manual charging to ensure continuous operation.
+The trained agent was deployed and evaluated on the prototype hardware without power generation in an indoor environment over a 9 day period with occasional manual charging to ensure continuous operation.
 This test evaluated the long-term performance of the agent and hardware system and the results are shown in Figure~\ref{fig:prototype_result}.
-The hardware did experience
+The hardware experienced a few outages with the longest one being 3.6 hours on 2024-10-07.
+The deployed agent performed on average 2.18 collections for every transmission action, which matches the training results in simulation.
 
@@
 
-The system was evaluated based on its power consumption, data transmission efficiency, and overall performance in real-world conditions. 
-The reinforcement learning algorithm optimized the power usage by balancing data collection, storage, and transmission to ensure continuous operation.
-
-Additionally, the power levels of the device over time were analyzed to assess energy efficiency.
-Figure~\ref{fig:power_levels} shows the variation in power consumption during different operational phases.
+Figure~\ref{fig:battery_vs_collection} shows the learned strategy for the agent under varying power levels.
+At full charge the agent prefers to altenate between collection and transmission, but as the power level decreases the agent prefers to collect data and store the messages for a median of 4 collections for every transmission at a power level of 30\%.
+At very low power levels the agent transmits data every time, which discharges the battery faster.
+This behaviour was an unexpected result of the simulation, where the device loses the messsages in the queue when power is lost.
+The agent learned to explot this inconsistency and did not want to lose a queue's worth of messages.
+The simuation will be corrected in future iterations.
 
@@
 In this study, we developed a reinforcement learning-enabled cattle tracker prototype designed to optimize power consumption while ensuring efficient data transmission and collection.
-The device successfully leverages reinforcement learning to dynamically adjust its operation based on available energy, enhancing battery life and operational efficiency.
+The device successfully leverages q-learning adjust its strategy based on available energy, time of day, and message queue length enhancing battery life and operational efficiency.
 By incorporating solar power and an intelligent decision-making framework, the prototype demonstrates a feasible approach to long-term, autonomous cattle tracking.
-Future work will focus on improving the efficiency of the reinforcement learning model by integrating real-time adaptation to environmental changes.
-Additional experiments in diverse field conditions will provide further validation of the system's robustness.
+Future work will focus on continue the iterative process of model development and hardware improvement to improve the system's robustness.
 Further refinements to the hardware, such as low-power components and custom PCB design, will enhance reliability and longevity.
-Moreover, expanding the cloud-based digital twin functionality can improve remote monitoring and predictive analytics for herd management.
 

### 2025-02-14 paper/main.tex
Add prototype diagram
Actual First Draft commit

@@
 Many farmers and researchers are interested in tracking movement of these cattle to asess the health of the herd, and to monitor the location of the cattle.
-Reasearchers are interested in behaviour of the herd like feeding, rumination, and resting and past research has shown that this behaviour can be correlated to sensor data from accelerometer \cite{unoldIoTBasedCowHealth2020}.
-Reasearchers would like a device that transmits information at high rates as higher fidelity information allows them to make better conclusions about the state of the herd.
+Reasearchers are interested feeding, rumination, and resting behaviour and past research has correlated it to sensor data from accelerometers \cite{unoldIoTBasedCowHealth2020}.
+Monitors benefit from tracking devices that transmits information at high rates as higher fidelity information allows them to make better conclusions about the state of the herd.
 These trackers take the form of collars, leg bands, or ear tags and are equipped with sensors, batteries, and communication modules.
 The installation of the trackers is a time consuming process so a tracker is expected to operate for extended periods of time (months) without human intervention.
-Given the size of the devices and the environment that they operate in, there is a limited power budget available to the device and it beocmes a balancing act to ensure contiunuous operation, transmission and data collection.
+Given the size of the devices and the environment that they operate in, there is a limited power budget to balance contiunuous operation, transmission, and data collection.
 
 The paradigm of Internet of Things (IoT) devices has emerged over the last two decades as a way to deploy sensors, communication modules, storage, compute, and intelligence throughout the environment.
-% The number of IoT devices is expected to grow to 75 billion by 2025 \cite{gubbiInternetThingsIoT2013}.
-The devices are often small, low power, and have limited communication capabilities, but provide insights into the environment that they are deployed in.
-Cattle trackers are an example of an IoT device that is deployed in the environment to provide insights into the health and location of the herd.
-These cattle trackers are strating to replace labour intensive manual observation with automated systems that can monitor the herd 24-7 \cite{unoldIoTBasedCowHealth2020} \cite{moutaouakilDigitalFarmingSurvey2023}.
+The devices are small, low power, and have limited communication capabilities, but provide insight into the environment that was previously not possible.
+These cattle trackers are IoT devices that replace labour intensive manual observation with automated systems that can monitor the herd 24-7 \cite{unoldIoTBasedCowHealth2020} \cite{moutaouakilDigitalFarmingSurvey2023}.
 IoT devices can operate as part of a network or as standalone units. 
@@
 
-Reinforcement learning is a branch of machine learning that allows an agent to learn optimum actions to take in an environment to maximize a reward.
-The agent learns these actions by interacting with the environment and receiving rewards for those actions.
+Reinforcement learning is a branch of machine learning that allows an agent to learn strategies to maximize a reward.
+The agent learns by interacting with the environment and receiving rewards for those actions.
 Reinforcement learning has been used in a variety of applications such as playing games, optimizing energy consumption, and optimizing the operation of IoT devices.
-The combination of reinforcement learning and IoT devices is an attractive option as it allows the device to make decisions without the need for human intervention.
-THe combvination of AI and IoT is sometimes called the Artificial intelligence of Things (AIoT)\cite{yamsaniIoTBasedLivestockMonitoring2024}, which includes a broad range of applications.
-The TinyML paradigm is a subset of AIoT and focuses on machine learning on devices with highly constrained resources, which often limits it to inference only\cite{rayReviewTinyMLStateoftheart2022}.
+The combination of reinforcement learning and IoT devices allows the device to make decisions without the need for human intervention.
+The combination of AI and IoT is called the Artificial intelligence of Things (AIoT)\cite{yamsaniIoTBasedLivestockMonitoring2024}, which includes a broad range of applications.
+The TinyML paradigm is a subset of AIoT and focuses on machine learning on devices with highly constrained resources and is often limited to inference only\cite{rayReviewTinyMLStateoftheart2022}.
 On one hand the addition of intelligence to the device allows it to make better decisions, but on the other hand the addition of intelligence to the device increases the power consumption of the device.
-Q-learning is a reinforcement learning algorithm that is used to optimize the actions that an agent takes in an environment and has been used in a variety of applications such as optimizing the operation of IoT devices.
 
 In this paper we present the development of a cattle tracker prototype capable of data collection for extended periods of time.
+The device is designed to alternate between data collection, data transmission, and sleeping to conserve power.
 It is not trivial to manually determine the optimal data collection strategy for the device under dynamic and complex conditions \cite{suttonReinforcementLearningIntroduction2020} so we propose the addition of reinforcement learning to automate the decision making process.
@@
 The IoT device likely fits other use cases, but the focus in this paper is on the cattle tracking application due to the requirements of collecting uninterrupted data, occasionally needing to check in, needing to operate for long periods of time, and the potential for a mobile solar panel, which results in drastically different power generation.
-
-The cattle tracker alternates between sleeping to conserve power and decision making. 
-It is faced with three decision points:
-
-\begin{itemize}
-  \item when to transmit data (high power consumption)
-  \item when to collect and store data (medium power consumption)
-  \item when to do nothing instead (lowest power consumption)
-\end{itemize}
-
-Data transmission is needed to check in and ensure that the device and the cattle is still operational, but consumes a large amount of power.
-Data collection and storage is needed to ensure uninterrupted data collection, and consumes a medium amount of power.
-Doing nothing consumes the least amount of power, but the device incurs gaps in the data collection, which should only be done to avoid draining the battery to an empty state.
-Doing nothing will get the device through the night/day to ensure that it doesn't end up in a weird state by discharging the battery completely.
 
@@
 
-%Cattle
-Agriculture and cattle tracking commonly employ IoT devices for monitoring. 
-In \cite{yamsaniIoTBasedLivestockMonitoring2024} the authors propose a network of IoT devices to monitor cattle, with edge devices for network communication and cloud services for data storage and analysis.
-Multiple standalone devices have also been developed for cattle tracking, with many focusing on the sensor output of the device.
-In \cite{duttaMOOnitorIoTBased2022} the authors present an IoT device that clasifies the behaviour of the cattle based on data from accelerometer, temperature via XGBoost and Random Forest classifiers.
-Their device lasts for an estimated 10 hours on a single charge.
-In \cite{unoldIoTBasedCowHealth2020} the authors present the CowMonitor an IoT device uses Bluetooth Low Energy (BLE) for communication with a hub and sends accelerometer and magnetometer data to the cloud.
-They transmit data without acknowledgement for oppurtunistic capture and lasts 5 years without battery replacement, but required the deployment of BLE hubs and WiFi ground stations thoughout the pasture.
+%Cattle and IoT
+Existing IoT-based cattle monitoring solutions have primarily focused on health tracking, behavior classification, and disease detection using machine learning and sensor networks. 
+Studies such as \cite{duttaMOOnitorIoTBased2022} have proposed multi-sensory IoT devices for cattle activity monitoring using XGBoost and Random Forest classifiers. 
+\cite{yamsaniIoTBasedLivestockMonitoring2024} introduced an ML-based livestock management system that classifies cattle behavior but does not optimize energy consumption. 
+Additionally, \cite{arshadFederatedLearningModel2024} and \cite{iRealTimeCattle2024} have applied LoRa-based IoT solutions and federated learning for disease detection, emphasizing data accuracy over power efficiency.
 
 % Reinforcement Learning
-The combination of intelligence and IoT devices allows a powerful solution to add intelligence to the edge.
-In \cite{hribarUsingDeepQLearning2019} the authors use deep Q-learning to extend the life of generic sensors networks with non-rechargable batteries. 
-The authors take advantage of corelation between sensor measurements to determine the rate at which sensors should transmit to maximize battery life.
+In contrast, our work integrates reinforcement learning to dynamically optimize power consumption in cattle trackers, ensuring long-term device operation while maintaining data quality. 
+Prior studies, such as \cite{hribarUsingDeepQLearning2019}, demonstrated that deep Q-learning could extend IoT device battery life, but they did not apply this approach to livestock monitoring. 
+Our system leverages RL-driven decision-making to balance data transmission, storage, and power conservation, addressing a key limitation in prior research and enhancing the sustainability of IoT-based livestock tracking solutions.
 
@@
 
-The design of an AIoT device is an iterative process, that loops between simulations, hardware development, measurement and redeployment.
+The design of an AIoT device is an iterative process, that contiunuously cycles between simulations, hardware development, measurement and redeployment.
 It is critical to have an automated process in place that can be updated without manual intervention.
-The use of reinforcement learning greatly simplifies the process and allows the device to adapt to new conditions without a manual update of collection strategies.
+The use of reinforcement learning greatly simplifies the process and allows the device to adapt to new conditions without the need for manual update of collection strategies.
 
@@
 The time of day informs the agent of the likelyhood of power generation and was discretized to 48 bins.
-The maximum power state was discretized into 100 power levels and the maximum number of messages was 5, forcing the agent to check in every 2.5 hours or otherwise risk losing messages.
+The maximum power state was discretized into 100 power levels and the maximum number of messages was 5, forcing the agent to check transmit every 2.5 hours or risk losing data.
 This resulted in 24,800 possible states for the agent to be in.
 
-The agent can choose one of three actions (i) transmit data (ii) collect and store data (iii) do nothing.
+The agent can choose one of three actions (i) do nothing (ii) collect and store data (iii) collect and transmit data.
 
@@
   \begin{array}{l}
-    a_0 = \text{``do nothing''}
-    a_1 = \text{``collect and store data''}, \\
-    a_2 = \text{``collect and transmit data''}, \\
+    a_0 = 0, \text{``do nothing''}
+    a_1 = 1, \text{``collect and store data''}, \\
+    a_2 = 2, \text{``collect and transmit data''}, \\
   \end{array}
@@
   \end{equation}
-Turning the action-value function into a $4800 \times 3$ table.
+Turning the action-value function into a $24800 \times 3$ table.
 
@@
 
-The goal of the agent is to maximize the number of messages sent to the cloud, while minimizing the power consumption of the device.
-It is not trivial to determine the reward values for the agent to maximize data collection.
-We consider the reward values to be hyperparameters that can be tuned to maximize the data collection and used a Bayesian optimization framework from the the python \verb|scikit-optimize| package to automatically tune the hyperparameters.
-The hyper parameters could be nagative and they were:
-\begin{itemize}
-  \item reward\_power\_loss: the reward for losing power
-  \item reward\_power\_multiplier: the reward for having excess power at every time step
-  \item reward\_action\_0: the reward for doing nothing
-  \item reward\_action\_1: the reward for collecting and storing data
-  \item reward\_action\_2: the reward for collecting and transmitting data
-  \item reward\_message\_count: the reward for the number of messages sent when transmission occurred
-\end{itemize}
+The goal of the agent is to maximize the number of messages sent to the cloud, while maximizing the power level of the device.
+It is not trivial to determine the reward values that maximize data collection.
+We consider the reward values to be hyperparameters that can be tuned to maximize the number of messages send and used a Bayesian optimization framework from the the python \verb|scikit-optimize| package to automatically tune the hyperparameters.
+The hyper parameters could be nagative and their values are shown in Table~\ref{tab:reward_parameters}.
+
+\begin{table}[h]
+  \centering
+  \begin{tabular}{l p{8cm} c}
+      \toprule
+      \textbf{Parameter} & \textbf{Description} & \textbf{Value} \\
+      \midrule
+      reward\_power\_loss & reward for losing power & 0.00  \\ 
+      reward\_power\_multiplier & reward per power level at every time step & 0.0001  \\ 
+      reward\_action\_0 & reward for doing nothing & -1.716 \\ 
+      reward\_action\_1 & reward for collecting and storing data & -0.600 \\ 
+      reward\_action\_2 & reward for collecting and transmitting data & -1.485 \\ 
+      reward\_message\_count & reward for each message at transmission & 0.841 \\ 
+      \bottomrule
+  \end{tabular}
+  \caption{Reward function parameters, their descriptions, and values extracted during hyperparameter tuning.}
+  \label{tab:reward_parameters}
+\end{table}
 
@@
 At every step, the environment would update the power level of the battery, the time of day, and the number of messages queued on the device depending on the action taken by the agent.
-The actual measured battery capacity was 1300 mAh and every 30 minutes the device would consume 5.0 mAh in sleeping.
-A power transmission would consume an additional 1.6 mAh and a data collection would consume an additional 0.1 mAh.
+The actual measured battery capacity was 1300 mAh (discretized into 100 power levels) and every 30 minutes the device would consume 5.0 mAh while sleeping.
+A message transmission action would consume an additional 1.6 mAh and a data collection action would consume an additional 0.1 mAh.
 
@@
 
-\begin{figure}[htbp]
+\begin{figure}[h]
   \centering
-  \includegraphics[width=0.5\textwidth]{./figs/prototype.png}
+  \begin{subfigure}{0.48\textwidth}
+      \centering
+      \includegraphics[width=0.75\linewidth]{./figs/prototype.png}
+      \caption{Prototype power level over a 9 day collection window}
+      \label{fig:prototype_real}
+  \end{subfigure}
+  \hfill
+  \begin{subfigure}{0.48\textwidth}
+      \centering
+      \includegraphics[width=\linewidth]{figs/prototype_diagram.png}
+      \caption{Overall architecture of the reinforcement learning-enabled cattle tracker}
+      \label{fig:prototype_diagram}
+  \end{subfigure}
   \caption{Prototype of the reinforcement learning-enabled cattle tracker.
@@
 \subsection{Firmware}
-The firmware for the prototype is written in C++ and is responsible for managing the power consumption of the device.
+The firmware for the prototype was Arduino C++ code and is responsible for collecting sensor data, making decisions, and transmitting data to the cloud (see Figure~\ref{fig:prototype_diagram}).
 The main loop runs at a 30 minute interval and the device alternates between sleeping and decision making.
@@
 \subsection{Cloud}
-The device transmits data to Azure IoT Central.
+The prototype transmits data to Azure IoT Hub, which interfaces with Azure IoT Central for dashboarding, and Blob storage for data retention.
 IoT Central is an attractive option as it allows the registering of the device for security purposes, the creation of dahsboards, and the creation of digital twins.
@@
 
-\subsection{Training Results}
-A baseline device that alternates between collection and transmission is used to compare the performance of the reinforcement learning agent.
-The results are shown in Table~\ref{tab:agent_comparison}.
+\subsection{Simulation Training Results}
+A baseline collection strategy that alternates between collection and transmission is used to compare the performance of the reinforcement learning agent.
+The baseline is set up to send data once per hour.
+The results comparing the trained agent and the baseline are shown in Table~\ref{tab:agent_comparison}.
 

### 2025-02-16 paper/main.tex
updated the main.tex and ref

@@
 
-There is an estimated 1.5 million beef cows in Alberta with a portion of those being free-range cattle who are allowed to graze on large pastures for extended periods of time.
-Many farmers and researchers are interested in tracking movement of these cattle to asess the health of the herd, and to monitor the location of the cattle.
-Reasearchers are interested feeding, rumination, and resting behaviour and past research has correlated it to sensor data from accelerometers \cite{unoldIoTBasedCowHealth2020}.
-Monitors benefit from tracking devices that transmits information at high rates as higher fidelity information allows them to make better conclusions about the state of the herd.
-These trackers take the form of collars, leg bands, or ear tags and are equipped with sensors, batteries, and communication modules.
-The installation of the trackers is a time consuming process so a tracker is expected to operate for extended periods of time (months) without human intervention.
-Given the size of the devices and the environment that they operate in, there is a limited power budget to balance contiunuous operation, transmission, and data collection.
-
-The paradigm of Internet of Things (IoT) devices has emerged over the last two decades as a way to deploy sensors, communication modules, storage, compute, and intelligence throughout the environment.
-The devices are small, low power, and have limited communication capabilities, but provide insight into the environment that was previously not possible.
-These cattle trackers are IoT devices that replace labour intensive manual observation with automated systems that can monitor the herd 24-7 \cite{unoldIoTBasedCowHealth2020} \cite{moutaouakilDigitalFarmingSurvey2023}.
-IoT devices can operate as part of a network or as standalone units. 
-For cattle tracking, a self-sufficient device that connects to the internet via cellular networks such as GSM, LTE, or even satellite services like Starlink is ideal, allowing farmers to monitor their herds remotely.
-However, these devices face power constraints due to the high data transmission rates required; solar panels can extend battery life, though their effectiveness depends on weather conditions and the cattle's orientation and location.
+Alberta is home to an estimated 1.5 million beef cattle, many of which are free-range and graze on large pastures for extended periods of time.
+Timely tracking of these animals is essential for farmers to assess herd health and monitor animal behaviors. 
+For now, various Internet of Things (IoT)-based solutions have been developed to replace labor-intensive manual observation with automated, real-time monitoring \cite{unoldIoTBasedCowHealth2020}.
+The devices are often small, low power, typically utilize collars, leg bands, or ear tags equipped with sensors and communication modules, and have limited communication capabilities, but provide insights into the environment that they are deployed in \cite{unoldIoTBasedCowHealth2020} \cite{moutaouakilDigitalFarmingSurvey2023}.
+However, their installation is a time consuming process, making long-term autonomous operation (months), without human intervention, a key requirement.
+Additionaly, power limitations remain another significant challenge, as frequent data transmission depletes battery, necessitrating the design of energy-efficient solutions.
+Solar panels can extend battery life, while their effectiveness relys heavily on weather conditions and the cattle's orientation and location.
 
@@
 The agent learns by interacting with the environment and receiving rewards for those actions.
-Reinforcement learning has been used in a variety of applications such as playing games, optimizing energy consumption, and optimizing the operation of IoT devices.
-The combination of reinforcement learning and IoT devices allows the device to make decisions without the need for human intervention.
-The combination of AI and IoT is called the Artificial intelligence of Things (AIoT)\cite{yamsaniIoTBasedLivestockMonitoring2024}, which includes a broad range of applications.
-The TinyML paradigm is a subset of AIoT and focuses on machine learning on devices with highly constrained resources and is often limited to inference only\cite{rayReviewTinyMLStateoftheart2022}.
-On one hand the addition of intelligence to the device allows it to make better decisions, but on the other hand the addition of intelligence to the device increases the power consumption of the device.
+The combination of reinforcement learning and IoT devices, also called the Artificial intelligence of Things (AIoT)\cite{yamsaniIoTBasedLivestockMonitoring2024}, offers a promising alternative by dynamically optimizing energy consumption through adaptive decision-making \cite{suttonReinforcementLearningIntroduction2020}.
+The TinyML paradigm, a subset of AIoT, focuses on machine learning on devices with highly constrained resources and is often limited to inference only\cite{rayReviewTinyMLStateoftheart2022}.
+Overall, the addition of intelligence to the device allows it to make better decisions, however, on the other hand the addition of intelligence to the device increases the power consumption of the device.
 
-In this paper we present the development of a cattle tracker prototype capable of data collection for extended periods of time.
-The device is designed to alternate between data collection, data transmission, and sleeping to conserve power.
-It is not trivial to manually determine the optimal data collection strategy for the device under dynamic and complex conditions \cite{suttonReinforcementLearningIntroduction2020} so we propose the addition of reinforcement learning to automate the decision making process.
-The main contribution of this paper is the development of a reinforcement learning agent that optimizes the power consumption of the device by allowing the agent to make decisions on when to transmit data, when to collect and store data, and when to do nothing.
-The hardware prototype uses solar power to extend the power budget of the device and the agent is trained to take advantage of excess power availability to increase the data transmission rate.
-The IoT device likely fits other use cases, but the focus in this paper is on the cattle tracking application due to the requirements of collecting uninterrupted data, occasionally needing to check in, needing to operate for long periods of time, and the potential for a mobile solar panel, which results in drastically different power generation.
+In this paper, we present a reinforcement learning-enabled cattle tracker that autonomously manages power consumption by deciding when to transmit, store, or delay data collection. 
+The device incorporates solar power to extend operational longevity, with the RL agent adapting to varying power availability. 
+Our approach ensures uninterrupted monitoring while balancing energy efficiency and data quality.
+By leveraging RL, the proposed system outperforms static scheduling approaches, allowing for the collection of uninterrupted data, as well as occasionally check in, also making it a viable solution for long-term autonomous cattle tracking.
 
@@
   \centering
+  \caption{Reward function parameters, their descriptions, and values extracted during hyperparameter tuning.}
   \begin{tabular}{l p{8cm} c}
@@
   \end{tabular}
-  \caption{Reward function parameters, their descriptions, and values extracted during hyperparameter tuning.}
   \label{tab:reward_parameters}
@@
 \subsection{Hardware}
-The hardware prototype is shown in \ref{fig:prototype} and consists of an Arduino microcontroller, a GPS device, a solar array, a power distribution system, and a battery.
+The hardware prototype is shown in figure \ref{fig:prototype} and consists of an Arduino microcontroller, a GPS device, a solar array, a power distribution system, and a battery.
 The detailed specifications are shown in table \ref{tab:hardware_inventory}.

### 2025-02-16 paper/main.tex
update the main.tex and ref

@@
   \centering
-  \caption{Reward function parameters, their descriptions, and values extracted during hyperparameter tuning.}
+  \caption{Reward function parameters, descriptions, and values extracted during hyperparameter tuning.}
   \begin{tabular}{l p{8cm} c}
@@
 
-\begin{table}[h!]
+\begin{table}[h]
   \centering
-  \caption{Hardware List for the Cattle Tracker Prototype}
-  \begin{tabular}{|l|l|l|}
-  \hline
-  \textbf{Hardware} & \textbf{Serial Number} &  \textbf{Description} \\ \hline
-  Arduino MKR WiFi 1010 & ABX00023 &  WiFi-enabled microcontroller board \\ \hline
-  Arduino MKR GPS Shield & ASX00017 &  Integrated GPS module \\ \hline
-  DFRobot Solar Power Manager & TPX00046 &  Manages and distributes solar cell and battery \\ \hline
-  Solar Panel & TPX00181 &  75x100x2.9 mm panel, 1W output \\ \hline
-  Sparkfun Fuel Gage & 1568-1273-ND & Analog sensor for power measurement \\ \hline
-  LiPo Battery & 2528 &  3.7V, 2000mAh capacity \\ \hline
+  \caption{Hardware list for the cattle tracker prototype}
+  \begin{tabular}{l p{8cm} c}
+      \toprule
+      \textbf{Hardware} & \textbf{Description} & \textbf{Serial Number} \\
+      \midrule
+      Arduino MKR WiFi 1010 & WiFi-enabled microcontroller board & ABX00023 \\
+      Arduino MKR GPS Shield & Integrated GPS module & ASX00017 \\
+      DFRobot Solar Power Manager & Manages and distributes solar cell and battery & TPX00046 \\
+      Solar Panel & 75x100x2.9 mm panel, 1W output & TPX00181 \\
+      Sparkfun Fuel Gage & Analog sensor for power measurement & 1568-1273-ND \\
+      LiPo Battery & 3.7V, 2000mAh capacity & 2528 \\
+      \bottomrule
   \end{tabular}
@@
   \centering
-  \begin{subfigure}{0.48\textwidth}
+  \begin{subfigure}[t]{0.48\textwidth}  % Align top
       \centering
-      \includegraphics[width=0.75\linewidth]{./figs/prototype.png}
-      \caption{Prototype power level over a 9 day collection window}
+      \includegraphics[height=6cm, keepaspectratio]{./figs/prototype.png}
+      \put(-170,160){\scriptsize \textbf{(A)}}  % Adjust position as needed
       \label{fig:prototype_real}
   \end{subfigure}
-  \hfill
-  \begin{subfigure}{0.48\textwidth}
+  \hspace{2mm}  % Reduce space between figures
+  \begin{subfigure}[t]{0.48\textwidth}  % Align top
       \centering
-      \includegraphics[width=\linewidth]{figs/prototype_diagram.png}
-      \caption{Overall architecture of the reinforcement learning-enabled cattle tracker}
+      \includegraphics[height=6cm, keepaspectratio]{./figs/prototype_diagram.png}
+      \put(-215,160){\scriptsize \textbf{(B)}}  % Adjust position as needed
       \label{fig:prototype_diagram}
@@
   \caption{Prototype of the reinforcement learning-enabled cattle tracker.
-The device consists of a solar panel for power harvesting, a rechargeable battery, and an embedded microcontroller for decision-making and communication.}
+(A) Physical prototype with hardware components.
+(B) System architecture diagram illustrating data processing and power management.}
   \label{fig:prototype}
@@
   \centering
+  \caption{Comparison of baseline agent and RL agent}
   \begin{tabular}{lcc}
@@
   \end{tabular}
-  \caption{Comparison of Baseline Agent and RL Agent}
   \label{tab:agent_comparison}

### 2025-02-16 paper/main.tex
Update main.text and references.bib

@@
 \subsection{Firmware}
-The firmware for the prototype was Arduino C++ code and is responsible for collecting sensor data, making decisions, and transmitting data to the cloud (see Figure~\ref{fig:prototype_diagram}).
+The firmware for the prototype was Arduino C++ code and is responsible for collecting sensor data, making decisions, and transmitting data to the cloud (see Figure~\ref{fig:prototype}).
 The main loop runs at a 30 minute interval and the device alternates between sleeping and decision making.
@@
       \includegraphics[width=\linewidth]{figs/prototype_power_level.png}
-      \caption{Prototype power level over a 9 day collection window}
+      \put(-195,138){\scriptsize \textbf{(A)}}  % Adjust position
       \label{fig:prototype_power}
@@
       \includegraphics[width=\linewidth]{figs/battery_level_vs_collection_ratio.png}
-      \caption{Learned agent strategies as a function of power levels}
+      \put(-195,138){\scriptsize \textbf{(B)}}  % Adjust position
       \label{fig:battery_vs_collection}
   \end{subfigure}
-  \caption{Collected data from the prototype hardware highlighting a nonlinear power level discharge and the learned agent strategies.}
+  \caption{Collected data from the prototype hardware. (A) Nonlinear power level discharge over a 9-day period. (B) Learned agent strategies adapting to varying power levels.}
   \label{fig:prototype_result}
@@
 
-Figure~\ref{fig:prototype_power} highlights the nonlinear discharge curve of the battery over the 9 day period. 
+Figure~\ref{fig:prototype_result}\subref{fig:battery_vs_collection}A highlights the nonlinear discharge curve of the battery over the 9 day period. 
 The discharge rate during sleeping is relatively constant and the nonlinearity is a result of the mapping function of voltage to remaining capacity.
@@
 
-Figure~\ref{fig:battery_vs_collection} shows the learned strategy for the agent under varying power levels.
+Figure~\ref{fig:prototype_result}\textsubscript{\subref{fig:battery_vs_collection}}B shows the learned strategy for the agent under varying power levels.
 At full charge the agent prefers to altenate between collection and transmission, but as the power level decreases the agent prefers to collect data and store the messages for a median of 4 collections for every transmission at a power level of 30\%.

### 2025-02-16 paper/main.tex
Update main.tex and references.bib
2nd Edit

@@
 In this study, we present the development of a reinforcement learning-enabled cattle tracker prototype designed to optimize power consumption while ensuring continuous data collection and transmission.
-The prototype integrates IoT sensors, solar power harvesting, and reinforcement learning-based decision-making to dynamically balance data transmission, storage, and energy conservation. 
+The prototype integrates IoT sensors, solar power collection, and reinforcement learning-based decision-making to dynamically balance data transmission, storage, and energy conservation. 
 The reinforcement learning agent autonomously determines the optimal transmission schedule based on power availability, maximizing data efficiency while prolonging battery life. 
@@
 
-Alberta is home to an estimated 1.5 million beef cattle, many of which are free-range and graze on large pastures for extended periods of time.
-Timely tracking of these animals is essential for farmers to assess herd health and monitor animal behaviors. 
-For now, various Internet of Things (IoT)-based solutions have been developed to replace labor-intensive manual observation with automated, real-time monitoring \cite{unoldIoTBasedCowHealth2020}.
-The devices are often small, low power, typically utilize collars, leg bands, or ear tags equipped with sensors and communication modules, and have limited communication capabilities, but provide insights into the environment that they are deployed in \cite{unoldIoTBasedCowHealth2020} \cite{moutaouakilDigitalFarmingSurvey2023}.
+Timely tracking of free range beef cattle is essential for farmers to assess herd health and monitor animal behaviors. 
+Internet of Things (IoT)-based solutions have been developed to replace labor-intensive manual observation with automated, real-time monitoring \cite{unoldIoTBasedCowHealth2020} \cite{yamsaniIoTBasedLivestockMonitoring2024}.
+The devices take the form of collars, leg bands, or ear tags and are small, low power, equipped with sensors and communication modules, and have limited communication capabilities, but provide insights into the environment that they are deployed in \cite{unoldIoTBasedCowHealth2020} \cite{moutaouakilDigitalFarmingSurvey2023}.
 However, their installation is a time consuming process, making long-term autonomous operation (months), without human intervention, a key requirement.
-Additionaly, power limitations remain another significant challenge, as frequent data transmission depletes battery, necessitrating the design of energy-efficient solutions.
+Power limitations remain another significant challenge, as frequent data transmission depletes battery, necessitrating the design of energy-efficient solutions.
 Solar panels can extend battery life, while their effectiveness relys heavily on weather conditions and the cattle's orientation and location.
 
-Reinforcement learning is a branch of machine learning that allows an agent to learn strategies to maximize a reward.
-The agent learns by interacting with the environment and receiving rewards for those actions.
-The combination of reinforcement learning and IoT devices, also called the Artificial intelligence of Things (AIoT)\cite{yamsaniIoTBasedLivestockMonitoring2024}, offers a promising alternative by dynamically optimizing energy consumption through adaptive decision-making \cite{suttonReinforcementLearningIntroduction2020}.
+Reinforcement learning (RL) is a branch of machine learning that allows an agent to learn strategies to maximize a reward.
+The agent learns by interacting with the environment and updating its behaviour based on the received rewards for those actions\cite{suttonReinforcementLearningIntroduction2020}.
+The combination of reinforcement learning and IoT devices, also called the Artificial intelligence of Things (AIoT), offers a promising alternative by dynamically optimizing energy consumption through adaptive decision-making.
 The TinyML paradigm, a subset of AIoT, focuses on machine learning on devices with highly constrained resources and is often limited to inference only\cite{rayReviewTinyMLStateoftheart2022}.
@@
 
-In this paper, we present a reinforcement learning-enabled cattle tracker that autonomously manages power consumption by deciding when to transmit, store, or delay data collection. 
-The device incorporates solar power to extend operational longevity, with the RL agent adapting to varying power availability. 
+In this paper, we present a reinforcement learning-enabled cattle tracker prototype that autonomously optimizes power consumption by deciding when to transmit, store, or delay data collection.
+The Q-learning-based agent adapting to varying power availability and the device incorporates solar power to extend operational longevity.
 Our approach ensures uninterrupted monitoring while balancing energy efficiency and data quality. 
-By leveraging RL, the proposed system outperforms static scheduling approaches, allowing for the collection of uninterrupted data, as well as occasionally check in, also making it a viable solution for long-term autonomous cattle tracking.
+By leveraging RL, the proposed system outperforms static scheduling approaches, allowing for the collection of uninterrupted data, as well as occasionally check in, making it a viable solution for long-term autonomous cattle tracking.
 
@@
 % IoT and Cloud
-The main challenges of IoT sensors are \cite{chenDeepReinforcementLearning2021} are (i) limited battery life, (ii) limited processing power, and (iii) limited storage.
-Devices that fall under the TinyML paradigm are often limited to inference only, as training a model on the device is computationally expensive.
-The addition of cloud communication opens up the possibility of training the model in the cloud and deploying a new model to the device that adapts to changes in the environment.
+IoT sensors are faced with (i) limited battery life, (ii) limited processing power, and (iii) limited storage\cite{chenDeepReinforcementLearning2021}.
+Model training is computationally expensive so devices that fall under the TinyML paradigm are often limited to inference only\cite{rayReviewTinyMLStateoftheart2022}.
+The addition of cloud communication opens the possibility of training the model in the cloud and deploying a new model to the device that adapts to changes in the environment.
 
@@
 % Reinforcement Learning
-In contrast, our work integrates reinforcement learning to dynamically optimize power consumption in cattle trackers, ensuring long-term device operation while maintaining data quality. 
-Prior studies, such as \cite{hribarUsingDeepQLearning2019}, demonstrated that deep Q-learning could extend IoT device battery life, but they did not apply this approach to livestock monitoring. 
-Our system leverages RL-driven decision-making to balance data transmission, storage, and power conservation, addressing a key limitation in prior research and enhancing the sustainability of IoT-based livestock tracking solutions.
+In contrast, our work integrates RL to dynamically optimize power consumption in cattle trackers, ensuring long-term device operation while maintaining data quality. 
+Prior studies, such as \cite{hribarUsingDeepQLearning2019}, demonstrated that deep Q-learning could extend IoT device battery life, but their work relies on a network of devices to determine whether correlated sensors should transmit data or save power.
+Our system leverages RL-driven decision-making for a single device to balance data transmission, storage, and power conservation, addressing a key limitation in prior research.
 
@@
 
-The design of an AIoT device is an iterative process, that contiunuously cycles between simulations, hardware development, measurement and redeployment.
-It is critical to have an automated process in place that can be updated without manual intervention.
-The use of reinforcement learning greatly simplifies the process and allows the device to adapt to new conditions without the need for manual update of collection strategies.
+AIoT device design is an iterative process involving simulations, hardware development, measurement, and redeployment
+Automation is essential to facilitate rapid development with reduced human intervention.
+RL helps this process by allowing the device to adapt to new parameters without the need for manual update of collection strategies.
 
@@
 
-Central to q-learning is the action-value function $Q(s,a)$, which is the expected cumulative reward of taking action $a$ in state $s$.
-The agent learns the optimal policy by updating the action-value function based on the reward that it receives over many training iterations according to:
+Central to Q-learning, a widely used RL approach, is the action-value function $Q(S_t, A_t)$, which is the expected cumulative reward of taking action $a$ in state $s$.
+The agent approximates the optimal policy by updating the action-value function based on the reward that it receives during training\cite{suttonReinforcementLearningIntroduction2020}:
 
 \begin{equation}
-  Q(s,a) \leftarrow Q(s,a) + \alpha \left( r + \gamma \max_{a'} Q(s',a') - Q(s,a) \right)
+Q^{new}(S_t, A_t) \leftarrow Q(S_t, A_t) + \alpha \left[ R_{t+1} + \gamma \max_{a} Q(S_{t+1}, a) - Q(S_t, A_t) \right]
 \end{equation}
-where $\alpha$ is the learning rate, $r$ is the reward, $\gamma$ is the discount factor, and $s'$ is the next state after taking action $a$ in state $s$.
 
-The state of the agent is a combination of the time of day $t$, the power level of the battery $p$, and the number of messages queued on the device $m$.
-The time of day informs the agent of the likelyhood of power generation and was discretized to 48 bins.
-The maximum power state was discretized into 100 power levels and the maximum number of messages was 5, forcing the agent to check transmit every 2.5 hours or risk losing data.
-This resulted in 24,800 possible states for the agent to be in.
+where $\alpha$ is the learning rate, $\gamma$ is the discount factor, and $R_{t+1}$ and $S_{t+1}$ is the reward and state after taking action $A_t$ in state $S_{t}$, respectively.
 
-The agent can choose one of three actions (i) do nothing (ii) collect and store data (iii) collect and transmit data.
+The agent's state consists of time of day $T$, battery power level $p$, and queued message count $m$.
+$T$ (discretized into 48 bins) informs the power generation likelihood, $p$ was discretized to 100 levels, and $m$ was discretized to 5 levels to enforce a transmission check every 2.5 hours to prevent data loss.
+This resulted in $T\times p \times m = 24,000$ possible states.
+The agent action space contains three discrete actions: $\mathcal{A}=\{a_0, a_1, a_2\}$, where $a_0=$delay data collection, $a_1=$collect and store data, and $a_2=$collect and transmit data, turning the action-value function into a $24000 \times 3$ table.
 
-\begin{equation}
-  \mathcal{A} = \left\{
-  \begin{array}{l}
-    a_0 = 0, \text{``do nothing''}
-    a_1 = 1, \text{``collect and store data''}, \\
-    a_2 = 2, \text{``collect and transmit data''}, \\
-  \end{array}
-  \right.
-  \end{equation}
-Turning the action-value function into a $24800 \times 3$ table.
+A tabular representation for $Q(S_t, A_t)$ was used as it turns the model inference into a table lookup with $O(1)$ complexity.
+Deep Q-learning alternatives, which use a neural network to approximate $Q(S_t, A_t)$ require additional on-device compuation and were not considered.
+A tabular representation does require larger memory or storage, but can be partially updated from the cloud to reflect new conditions.
 
-A tabular representation of the action-value function $Q(a,s)$ was used as it turns the model inference into a table lookup with $O(1)$ complexity.
-Alternatives to the tabular representation include deep Q-learning, which uses a neural network to approximate the action-value function, and is more computationally expensive.
-A tabular representation only requires a sufficiently large memory to store the table.
-An additional benefit is that the table can be easily updated from the cloud as individual entries can be updated to reflect new conditions.
-
-The goal of the agent is to maximize the number of messages sent to the cloud, while maximizing the power level of the device.
-It is not trivial to determine the reward values that maximize data collection.
-We consider the reward values to be hyperparameters that can be tuned to maximize the number of messages send and used a Bayesian optimization framework from the the python \verb|scikit-optimize| package to automatically tune the hyperparameters.
-The hyper parameters could be nagative and their values are shown in Table~\ref{tab:reward_parameters}.
+The goal of the agent is to maximize the number of messages sent to the cloud and the power level of the device.
+It is not trivial to determine the reward values that achieve this goal.
+We consider the reward values to be hyperparameters and used a Bayesian optimization framework from the the python \verb|scikit-optimize| package to tune them.
+The hyperparameters (Table~\ref{tab:reward_parameters}) show that the most negative reward is for delaying collection. 
+The reward for transmitting data also collects the reward per message, which only becomes attractive if there are messages in the queue.
 
@@
   \centering
-  \caption{Reward function parameters, descriptions, and values extracted during hyperparameter tuning.}
+  \caption{Reward function parameters extracted during hyperparameter tuning.}
   \begin{tabular}{l p{8cm} c}
@@
       \midrule
-      reward\_power\_loss & reward for losing power & 0.00  \\ 
+      reward\_power\_loss & reward for full discharge of battery & 0.00  \\ 
       reward\_power\_multiplier & reward per power level at every time step & 0.0001  \\ 
-      reward\_action\_0 & reward for doing nothing & -1.716 \\ 
+      reward\_action\_0 & reward for delaying collection & -1.716 \\ 
       reward\_action\_1 & reward for collecting and storing data & -0.600 \\ 
@@
 
-The agent was trained on a simulated environment with the power consumption and generation numbers from the prototype.
-At every step, the environment would update the power level of the battery, the time of day, and the number of messages queued on the device depending on the action taken by the agent.
-The actual measured battery capacity was 1300 mAh (discretized into 100 power levels) and every 30 minutes the device would consume 5.0 mAh while sleeping.
-A message transmission action would consume an additional 1.6 mAh and a data collection action would consume an additional 0.1 mAh.
+The agent was trained in a simulated environment with the power consumption and generation numbers collected from the prototype hardware.
+At each step, the environment updates the battery level, time of day, and queued messages based on the agent's action.
+The 1300 mAh battery (discretized into 100 levels) drains 5.0 mAh per 30 minutes in sleep mode, while transmission and data collection consume 1.6 mAh and 0.1 mAh, respectively.
 
-Sample solar gain data from 17 days in June in Edmonton was turned into an average sunny day and an average cloudy day. 
-The environment would switch between the two days to simulate the weather conditions with an 80\% chance of a cloudy day.
-These conditions ensured that there was not sufficient power available to transmit data at every time step and forced the agent to learn strategies to increase message count.
-
-The training run was repeated 4 times each with a different random seed at every iteration of the hyper parameter search.
-Once an optimum set of hyperparameters was found, the action-value function was deployed into an h-file that was used in the firmware of the device.
-This automated approach was less labour intensive than manual tuning and has shown to be effective in optimizing the reward values.
+Solar gain data from 17 June days in Edmonton was averaged into sunny and cloudy profiles, with an 80\% probability of a cloudy day. 
+This constraint limited power availability, forcing the agent to optimize message transmission. 
+The training process was repeated four times with different seeds for each hyperparameter search iteration. 
+Once the optimal hyperparameters were found, the action-value function was exported as an h-file for firmware deployment. 
+This automated approach reduced manual effort and effectively optimized reward values.
 
@@
 The detailed specifications are shown in table \ref{tab:hardware_inventory}. 
+Prototype testing revealed high sleep power consumption (9.9 mA) due to always-on LEDs and integrated components. 
+A custom PCB with full component isolation is planned to reduce sleep current to $\mu A$ levels.
+An unexpected benefit of high power draw was a faster development cycle, as battery depletion occurred in days rather than weeks.
 
@@
   \caption{Hardware list for the cattle tracker prototype}
-  \begin{tabular}{l p{8cm} c}
+  \begin{tabular}{l p{6.5cm} c}
       \toprule
@@
       Arduino MKR GPS Shield & Integrated GPS module & ASX00017 \\
-      DFRobot Solar Power Manager & Manages and distributes solar cell and battery & TPX00046 \\
+      DFRobot Solar Power Manager & Power dsitribution board & TPX00046 \\
       Solar Panel & 75x100x2.9 mm panel, 1W output & TPX00181 \\
@@
 \end{table}
-
-Testing on prototype-level hardware showed a high power consumption of 9.9 mA during sleeping due to the presence of multiple LEDs and integrated components that could not be disabled. 
-A custom-build PCB with isolation of every component should experience a sleep current on the order of $\mu A$ and is planned for the the next iteration hardware to reduce power consumption during sleep.
-An unexpected side-benefit of a prototype with large power consumption is a fast development feedback cycle where the prototype drains the battery in days, rather than weeks.
 
@@
   \end{subfigure}
-  \caption{Prototype of the reinforcement learning-enabled cattle tracker.
-(A) Physical prototype with hardware components.
-(B) System architecture diagram illustrating data processing and power management.}
+  \caption{(A) Physical prototype with hardware components. (B) System architecture diagram illustrating data processing and power management.}
   \label{fig:prototype}
@@
 \subsection{Firmware}
-The firmware for the prototype was Arduino C++ code and is responsible for collecting sensor data, making decisions, and transmitting data to the cloud (see Figure~\ref{fig:prototype}).
-The main loop runs at a 30 minute interval and the device alternates between sleeping and decision making.
-At each decision point the device reads the time, the power level of the battery, and the number of messages stored in the device and retrivees a recommended action.
-The device then takes the recommended action and updates the power level of the battery.
-
-The device is limited in its memory to store the policy.
-To pair down the table, only the highest action for each state was selected.
-The policy was then stored as values that were bit-shifted to fit into a 32-bit integer.
+The prototype firmware, written in Arduino C++, handles sensor data collection, decision-making, and cloud transmission (see Figure~\ref{fig:prototype} B). 
+The main loop runs every 30 minutes, alternating between sleep mode and decision-making.
+At each decision point, the device reads the time, battery level, and queued messages, retrieves a recommended action, and executes it.
+Due to memory constraints, only the highest-value action per state was stored, reducing the policy table size. 
+The policy was bit-shifted into and array of 32-bit integers for efficient storage in memory.
 
 \subsection{Cloud}
-The prototype transmits data to Azure IoT Hub, which interfaces with Azure IoT Central for dashboarding, and Blob storage for data retention.
-IoT Central is an attractive option as it allows the registering of the device for security purposes, the creation of dahsboards, and the creation of digital twins.
-The digital twin is a digital representation of the device and can be used to monitor the device and update the state of the device.
-The cloud enables two-way communication, which allows the device to receive updates and new policies from the cloud whenever it is connected.
-For example the cloud can update the device with a new policy for a new season, or a new policy for a new location that affects the solar gain.
+The prototype transmits data to Azure IoT Hub, which integrates with IoT Central for dashboarding and Blob Storage for data retention (see Figure~\ref{fig:prototype} B). 
+IoT Central supports device registration, dashboard creation, and digital twins, enabling remote monitoring and state updates.
+The cloud facilitates two-way communication, allowing the device to receive policy updates for potential seasonal changes or location-specific solar gain variations.
 
 \section{Results}
-The developed reinforcement learning-enabled cattle tracker prototype was successfully assembled and tested. 
-The prototype consists of key components such as a solar panel, battery, communication module, and processing unit, as shown in Figure~\ref{fig:prototype}.
 
 \subsection{Simulation Training Results}
-A baseline collection strategy that alternates between collection and transmission is used to compare the performance of the reinforcement learning agent.
-The baseline is set up to send data once per hour.
-The results comparing the trained agent and the baseline are shown in Table~\ref{tab:agent_comparison}.
+To evaluate the effectiveness of the reinforcement learning agent, we compare it to a baseline collection strategy that alternates between data collection and transmission, sending data once per hour.
+This represents a simple, fixed scheduling approach.
+Table~\ref{tab:agent_comparison} shows that the RL agent reduces transmissions, conserving power while increasing collection actions, leading to a higher message count (+4.3\%) and a more sustainable average power level (76 vs.
+34).
+The trained agent optimized collection-to-transmission behavior, performing 2.12 collection actions per transmission, compared to the 1:1 fixed schedule of the baseline. 
+This suggests the agent learned an adaptive strategy, balancing energy efficiency with message throughput.
 
@@
 
-The reinforcement learning agent outperforms the baseline agent by increasing the number of messages sent to the cloud by 4.3\% while maintaining a higher average power level by avoiding excess transmission actions.
-The trained agent performed on average 2.12 collection actions for every transmission action, compared to a 1:1 ratio for the baseline agent.
-
 \subsection{Deployment Results}
-
-The trained agent was deployed and evaluated on the prototype hardware without power generation in an indoor environment over a 9 day period with occasional manual charging to ensure continuous operation.
-This test evaluated the long-term performance of the agent and hardware system and the results are shown in Figure~\ref{fig:prototype_result}.
-The hardware experienced a few outages with the longest one being 3.6 hours on 2024-10-07.
-The deployed agent performed on average 2.18 collections for every transmission action, which matches the training results in simulation.
+The trained agent was deployed on the prototype hardware in an indoor environment without power generation for a 9-day evaluation period, with occasional manual charging to ensure continuous operation.
+This test assessed the agent’s long-term decision-making performance in real hardware and validated its ability to operate under realistic power constraints.
+The results are shown in Figure~\ref{fig:prototype_result}.
+During the test, the hardware experienced several outages, the longest lasting 3.6 hours on 2024-10-07.  
+The agent maintained an average collection-to-transmission ratio of 2.18, which closely matches the simulation results.
 
@@
   \end{subfigure}
-  \caption{Collected data from the prototype hardware. (A) Nonlinear power level discharge over a 9-day period. (B) Learned agent strategies adapting to varying power levels.}
+  \caption{(A) Nonlinear power level discharge over a 9-day period. (B) Learned agent strategy adapting to varying power levels.}
   \label{fig:prototype_result}
@@
 
-Figure~\ref{fig:prototype_result}\subref{fig:battery_vs_collection}A highlights the nonlinear discharge curve of the battery over the 9 day period. 
+Figure~\ref{fig:prototype_result}\subref{fig:battery_vs_collection}A illustrates the nonlinear battery discharge curve observed over the 9-day period.  
 The discharge rate during sleeping is relatively constant and the nonlinearity is a result of the mapping function of voltage to remaining capacity.
-This nonlinearity poses a challenge to the agent, since it increases the error in future reward and remaining capacity estimation at lower power levels.
+While sleep mode consumption remains constant, the voltage-to-capacity mapping introduces nonlinearity, making it difficult for the agent to accurately predict future battery levels at low charge states.
 
-Figure~\ref{fig:prototype_result}\textsubscript{\subref{fig:battery_vs_collection}}B shows the learned strategy for the agent under varying power levels.
-At full charge the agent prefers to altenate between collection and transmission, but as the power level decreases the agent prefers to collect data and store the messages for a median of 4 collections for every transmission at a power level of 30\%.
-At very low power levels the agent transmits data every time, which discharges the battery faster.
-This behaviour was an unexpected result of the simulation, where the device loses the messsages in the queue when power is lost.
-The agent learned to explot this inconsistency and did not want to lose a queue's worth of messages.
-The simuation will be corrected in future iterations.
+Figure~\ref{fig:prototype_result}\subref{fig:battery_vs_collection}B shows the agent's adaptive strategy under varying power levels.
+The agent alternates between collection and transmission at full charge, prioritizes storage with a 4:1 collection-to-transmission ratio at 30\% power, and transmits every collected message at low power which accelerates battery drain.
+Messages are stored in volatile memory so the queue is lost when the battery is fully discharged.
+The agent learned to transmit every time to prevent data loss at low power levels.
+This revealed a short-coming of the prototype and future iterations will store the queue in non-volatile memory to prevent data loss.
 
 \section{Conclusion}
-In this study, we developed a reinforcement learning-enabled cattle tracker prototype designed to optimize power consumption while ensuring efficient data transmission and collection.
-The device successfully leverages q-learning adjust its strategy based on available energy, time of day, and message queue length enhancing battery life and operational efficiency.
-By incorporating solar power and an intelligent decision-making framework, the prototype demonstrates a feasible approach to long-term, autonomous cattle tracking.
-Future work will focus on continue the iterative process of model development and hardware improvement to improve the system's robustness.
-Further refinements to the hardware, such as low-power components and custom PCB design, will enhance reliability and longevity.
+In this study, we developed a Q-learning-enabled cattle tracker that optimizes power consumption while ensuring efficient data collection and transmission. 
+The device dynamically adjusts its strategy based on energy availability, time of day, and message queue length, improving battery life and efficiency. 
+By integrating solar power with adaptive decision-making, the prototype offers a viable solution for long-term autonomous tracking. 
+Future work will focus on hardware optimizations, including low-power components and a custom PCB to enhance reliability and longevity. 
+Additionally, we aim to increase the message queue length, improve the reward function to account for the age of the messages, and incorporate non-volatile memory storage.
 

### 2025-02-17 paper/main.tex
First pre-submission draft

@@
 
-\title{Development of a Reinforcement Learning Enabled Cattle Tracker Prototype}
+\title{Power-Aware Data Collection for IoT Cattle Tracking Using Reinforcement Learning}
 \maketitle
@@
 \begin{abstract}
-The increasing demand for precision livestock monitoring has led to the development of Internet of Things (IoT)-enabled solutions for real-time tracking and health assessment of cattle. 
-In this study, we present the development of a reinforcement learning-enabled cattle tracker prototype designed to optimize power consumption while ensuring continuous data collection and transmission.
-The prototype integrates IoT sensors, solar power collection, and reinforcement learning-based decision-making to dynamically balance data transmission, storage, and energy conservation. 
-The reinforcement learning agent autonomously determines the optimal transmission schedule based on power availability, maximizing data efficiency while prolonging battery life. 
-The agent is trained on measured prototype data and the reward function is tuned from a Bayesian optimization to maximize the message collection.
-Experimental evaluations demonstrate the prototype's capability to sustain long-term operation under real-world environmental constraints. 
-The results highlight the feasibility of integrating machine learning with IoT devices for adaptive, energy-efficient cattle monitoring.
-Our findings contribute to the broader field of smart agriculture by demonstrating how reinforcement learning can be leveraged to enhance IoT-based livestock management. 
-Future work will focus on optimizing the decision-making model for diverse environmental conditions and scaling the system for larger deployments.
+IoT-based cattle trackers enable real-time livestock monitoring, but their energy constraints limit long-term deployment in remote environments.
+Efficient power management is critical to ensure continuous data collection while maximizing battery life.
+This study presents a Q-learning-enabled IoT device that dynamically adjusts data collection and transmission schedules to optimize power efficiency.
+The reinforcement learning agent adapts its strategy based on battery state, time of day, and message queue size.
+The reward function is fine-tuned using Bayesian optimization to maximize data continuity and power levels.
+Simulation results show an increase in 4.3\% in message count at a substantially higher average power level compared to a fixed scheduling approach of alternating collection and transmission.
+The agent was deployed on prototype hardware over 9 days and demonstrated the ability to adaptively collect and transmit data to the cloud. 
+These findings highlight the potential for power-adaptive reinforcement learning in IoT systems, contributing to energy-efficient smart agriculture.
+Future work will focus on hardware hardening and the expansion of the agent to account for message age.
 
@@
 \section{Introduction}
-
-Timely tracking of free range beef cattle is essential for farmers to assess herd health and monitor animal behaviors. 
+Timely tracking of free-range beef cattle is essential to assess herd health and monitor animal rest, grazing, and rumination behaviours.
+Understanding grazing patterns on pasture enables cattle producers to optimize feed efficiency, reducing production cost, and lowering environmental impact.
 Internet of Things (IoT)-based solutions have been developed to replace labor-intensive manual observation with automated, real-time monitoring \cite{unoldIoTBasedCowHealth2020} \cite{yamsaniIoTBasedLivestockMonitoring2024}.

### 2025-02-17 paper/main.tex
Submission Draft for Paper

@@
 The reinforcement learning agent adapts its strategy based on battery state, time of day, and message queue size.
-The reward function is fine-tuned using Bayesian optimization to maximize data continuity and power levels.
+The reward function is tuned using Bayesian optimization to maximize data continuity and power levels.
 Simulation results show an increase in 4.3\% in message count at a substantially higher average power level compared to a fixed scheduling approach of alternating collection and transmission.
@@
 \section{Introduction}
-Timely tracking of free-range beef cattle is essential to assess herd health and monitor animal rest, grazing, and rumination behaviours.
+Timely tracking of free-range beef cattle is essential to assess herd health and monitor rest, grazing, and rumination behaviours.
 Understanding grazing patterns on pasture enables cattle producers to optimize feed efficiency, reducing production cost, and lowering environmental impact.
 Internet of Things (IoT)-based solutions have been developed to replace labor-intensive manual observation with automated, real-time monitoring \cite{unoldIoTBasedCowHealth2020} \cite{yamsaniIoTBasedLivestockMonitoring2024}.
-The devices take the form of collars, leg bands, or ear tags and are small, low power, equipped with sensors and communication modules, and have limited communication capabilities, but provide insights into the environment that they are deployed in \cite{unoldIoTBasedCowHealth2020} \cite{moutaouakilDigitalFarmingSurvey2023}.
+The devices take the form of collars, leg bands, or ear tags and are small, low power, and equipped with sensors and communication modules.
+While their communication abilities are limited, they provide valuable insights into the environment that they are deployed in \cite{unoldIoTBasedCowHealth2020} \cite{moutaouakilDigitalFarmingSurvey2023}.
 However, their installation is a time consuming process, making long-term autonomous operation (months), without human intervention, a key requirement.
-Power limitations remain another significant challenge, as frequent data transmission depletes battery, necessitrating the design of energy-efficient solutions.
-Solar panels can extend battery life, while their effectiveness relys heavily on weather conditions and the cattle's orientation and location.
+Power limitations remain another significant challenge, as frequent data transmission depletes battery, necessitating the design of energy-efficient solutions.
+Solar panels can extend battery life but their effectiveness relies heavily on weather conditions and the cattle's orientation and location.
 
@@
 
-In this paper, we present a reinforcement learning-enabled cattle tracker prototype that autonomously optimizes power consumption by deciding when to transmit, store, or delay data collection.
-The Q-learning-based agent adapting to varying power availability and the device incorporates solar power to extend operational longevity.
-Our approach ensures uninterrupted monitoring while balancing energy efficiency and data quality. 
-By leveraging RL, the proposed system outperforms static scheduling approaches, allowing for the collection of uninterrupted data, as well as occasionally check in, making it a viable solution for long-term autonomous cattle tracking.
+In this paper, we present a reinforcement learning-enabled cattle tracker prototype that optimizes power consumption by deciding when to transmit, store, or delay data collection.
+The physical device incorporates a solar panel to extend operational longevity, while the Q-learning-based agent adapts to varying power availability to maintain continuous data collection. 
+Our approach leverages RL to outperform static scheduling approaches, ensuring uninterrupted monitoring while strategically checking in and transmitting data, making it a viable solution for long-term autonomous cattle tracking.
 
@@
 Automation is essential to facilitate rapid development with reduced human intervention.
-RL helps this process by allowing the device to adapt to new parameters without the need for manual update of collection strategies.
+RL helps this process by allowing the agent to adapt to new device properties without the need for manual update of collection strategies.
 
@@
 
-Central to Q-learning, a widely used RL approach, is the action-value function $Q(S_t, A_t)$, which is the expected cumulative reward of taking action $a$ in state $s$.
+Central to Q-learning, a widely used RL approach, is the action-value function $Q(S_t, A_t)$, which is the expected cumulative reward of taking action $A_t$ in state $S_t$.
 The agent approximates the optimal policy by updating the action-value function based on the reward that it receives during training\cite{suttonReinforcementLearningIntroduction2020}:
@@
 
-where $\alpha$ is the learning rate, $\gamma$ is the discount factor, and $R_{t+1}$ and $S_{t+1}$ is the reward and state after taking action $A_t$ in state $S_{t}$, respectively.
+where $\alpha$ is the learning rate, $\gamma$ is the discount factor, and $R_{t+1}$ and $S_{t+1}$ are the reward and state after taking action $A_t$ in state $S_{t}$, respectively.
 
@@
 A tabular representation for $Q(S_t, A_t)$ was used as it turns the model inference into a table lookup with $O(1)$ complexity.
-Deep Q-learning alternatives, which use a neural network to approximate $Q(S_t, A_t)$ require additional on-device compuation and were not considered.
+Deep Q-learning alternatives, which use a neural network to approximate $Q(S_t, A_t)$ require additional on-device computation and were not considered.
 A tabular representation does require larger memory or storage, but can be partially updated from the cloud to reflect new conditions.
@@
 It is not trivial to determine the reward values that achieve this goal.
-We consider the reward values to be hyperparameters and used a Bayesian optimization framework from the the python \verb|scikit-optimize| package to tune them.
+We consider the reward values to be hyperparameters and used a Bayesian optimization framework from the python \verb|scikit-optimize| package to tune them.
 The hyperparameters (Table~\ref{tab:reward_parameters}) show that the most negative reward is for delaying collection. 
@@
 At each step, the environment updates the battery level, time of day, and queued messages based on the agent's action.
-The 1300 mAh battery (discretized into 100 levels) drains 5.0 mAh per 30 minutes in sleep mode, while transmission and data collection consume 1.6 mAh and 0.1 mAh, respectively.
+The battery (discretized into 100 levels) drains 5.0 mAh per 30 minutes in sleep mode, while transmission and data collection consume 1.6 mAh and 0.1 mAh, respectively.
 
@@
 \subsection{Hardware}
-The hardware prototype is shown in figure \ref{fig:prototype} and consists of an Arduino microcontroller, a GPS device, a solar array, a power distribution system, and a battery.
-The detailed specifications are shown in table \ref{tab:hardware_inventory}. 
-Prototype testing revealed high sleep power consumption (9.9 mA) due to always-on LEDs and integrated components. 
+The hardware prototype is shown in Figure \ref{fig:prototype} and consists of an Arduino microcontroller, a GPS device, a solar array, a power distribution system, and a battery.
+The detailed specifications are shown in Table \ref{tab:hardware_inventory}. 
+Prototype testing revealed high sleep power consumption (9.9 mA) caused by always-on LEDs and integrated components. 
 A custom PCB with full component isolation is planned to reduce sleep current to $\mu A$ levels.
@@
       Arduino MKR GPS Shield & Integrated GPS module & ASX00017 \\
-      DFRobot Solar Power Manager & Power dsitribution board & TPX00046 \\
+      DFRobot Solar Power Manager & Power distribution board & DFR0559 \\
       Solar Panel & 75x100x2.9 mm panel, 1W output & TPX00181 \\
-      Sparkfun Fuel Gage & Analog sensor for power measurement & 1568-1273-ND \\
-      LiPo Battery & 3.7V, 2000mAh capacity & 2528 \\
+      Sparkfun Fuel Gage & Power measurement sensor & MAX17043 \\
+      LiPo Battery & 3.7V, 2100mAh rated, 1300mAh eff. capacity & 2528 \\
       \bottomrule
@@
 The main loop runs every 30 minutes, alternating between sleep mode and decision-making.
-At each decision point, the device reads the time, battery level, and queued messages, retrieves a recommended action, and executes it.
+At each decision point, the device reads the time, battery level, and queued messages, retrieves a recommended action from the policy table, and executes it.
 Due to memory constraints, only the highest-value action per state was stored, reducing the policy table size. 
-The policy was bit-shifted into and array of 32-bit integers for efficient storage in memory.
+The policy was bit-shifted into an array of 32-bit integers for efficient storage in memory.
 

### 2025-02-19 paper/appendix.tex
Simple Power Drain Baseline Update
- Create a test that looks at simple solar drain baselines and compares it across 5 different possible strategies
- Update typo which affects the paper slightly
- Add an appendix.tex file to keep track of solid experiments that we have performed
- rename DQN agent to tabular_agent since that is more accurate what it is.

--- 2025-02-17 paper/main.tex
@@
 
-\title{Power-Aware Data Collection for IoT Cattle Tracking Using Reinforcement Learning}
+\title{Eco Tracker Appendix}
 \maketitle
@@
 \begin{abstract}
-IoT-based cattle trackers enable real-time livestock monitoring, but their energy constraints limit long-term deployment in remote environments.
-Efficient power management is critical to ensure continuous data collection while maximizing battery life.
-This study presents a Q-learning-enabled IoT device that dynamically adjusts data collection and transmission schedules to optimize power efficiency.
-The reinforcement learning agent adapts its strategy based on battery state, time of day, and message queue size.
-The reward function is tuned using Bayesian optimization to maximize data continuity and power levels.
-Simulation results show an increase in 4.3\% in message count at a substantially higher average power level compared to a fixed scheduling approach of alternating collection and transmission.
-The agent was deployed on prototype hardware over 9 days and demonstrated the ability to adaptively collect and transmit data to the cloud. 
-These findings highlight the potential for power-adaptive reinforcement learning in IoT systems, contributing to energy-efficient smart agriculture.
-Future work will focus on hardware hardening and the expansion of the agent to account for message age.
+This is an appendix of additional material that we don't have room to put into the paper.
+This material will allow us to expand if needed and is here to test our own understanding of the the underlying mechanisms.
 
@@
 
-\section{Introduction}
-Timely tracking of free-range beef cattle is essential to assess herd health and monitor rest, grazing, and rumination behaviours.
-Understanding grazing patterns on pasture enables cattle producers to optimize feed efficiency, reducing production cost, and lowering environmental impact.
-Internet of Things (IoT)-based solutions have been developed to replace labor-intensive manual observation with automated, real-time monitoring \cite{unoldIoTBasedCowHealth2020} \cite{yamsaniIoTBasedLivestockMonitoring2024}.
-The devices take the form of collars, leg bands, or ear tags and are small, low power, and equipped with sensors and communication modules.
-While their communication abilities are limited, they provide valuable insights into the environment that they are deployed in \cite{unoldIoTBasedCowHealth2020} \cite{moutaouakilDigitalFarmingSurvey2023}.
-However, their installation is a time consuming process, making long-term autonomous operation (months), without human intervention, a key requirement.
-Power limitations remain another significant challenge, as frequent data transmission depletes battery, necessitating the design of energy-efficient solutions.
-Solar panels can extend battery life but their effectiveness relies heavily on weather conditions and the cattle's orientation and location.
-
-Reinforcement learning (RL) is a branch of machine learning that allows an agent to learn strategies to maximize a reward.
-The agent learns by interacting with the environment and updating its behaviour based on the received rewards for those actions\cite{suttonReinforcementLearningIntroduction2020}.
-The combination of reinforcement learning and IoT devices, also called the Artificial intelligence of Things (AIoT), offers a promising alternative by dynamically optimizing energy consumption through adaptive decision-making.
-The TinyML paradigm, a subset of AIoT, focuses on machine learning on devices with highly constrained resources and is often limited to inference only\cite{rayReviewTinyMLStateoftheart2022}.
-Overall, the addition of intelligence to the device allows it to make better decisions, however, on the other hand the addition of intelligence to the device increases the power consumption of the device.
-
-In this paper, we present a reinforcement learning-enabled cattle tracker prototype that optimizes power consumption by deciding when to transmit, store, or delay data collection.
-The physical device incorporates a solar panel to extend operational longevity, while the Q-learning-based agent adapts to varying power availability to maintain continuous data collection. 
-Our approach leverages RL to outperform static scheduling approaches, ensuring uninterrupted monitoring while strategically checking in and transmitting data, making it a viable solution for long-term autonomous cattle tracking.
-
-\section{Related Work}
-% IoT and Cloud
-IoT sensors are faced with (i) limited battery life, (ii) limited processing power, and (iii) limited storage\cite{chenDeepReinforcementLearning2021}.
-Model training is computationally expensive so devices that fall under the TinyML paradigm are often limited to inference only\cite{rayReviewTinyMLStateoftheart2022}.
-The addition of cloud communication opens the possibility of training the model in the cloud and deploying a new model to the device that adapts to changes in the environment.
-
-%Cattle and IoT
-Existing IoT-based cattle monitoring solutions have primarily focused on health tracking, behavior classification, and disease detection using machine learning and sensor networks. 
-Studies such as \cite{duttaMOOnitorIoTBased2022} have proposed multi-sensory IoT devices for cattle activity monitoring using XGBoost and Random Forest classifiers. 
-\cite{yamsaniIoTBasedLivestockMonitoring2024} introduced an ML-based livestock management system that classifies cattle behavior but does not optimize energy consumption. 
-Additionally, \cite{arshadFederatedLearningModel2024} and \cite{iRealTimeCattle2024} have applied LoRa-based IoT solutions and federated learning for disease detection, emphasizing data accuracy over power efficiency.
-
-% Reinforcement Learning
-In contrast, our work integrates RL to dynamically optimize power consumption in cattle trackers, ensuring long-term device operation while maintaining data quality. 
-Prior studies, such as \cite{hribarUsingDeepQLearning2019}, demonstrated that deep Q-learning could extend IoT device battery life, but their work relies on a network of devices to determine whether correlated sensors should transmit data or save power.
-Our system leverages RL-driven decision-making for a single device to balance data transmission, storage, and power conservation, addressing a key limitation in prior research.
-
-\section{System Design and Architecture}
-
-AIoT device design is an iterative process involving simulations, hardware development, measurement, and redeployment
-Automation is essential to facilitate rapid development with reduced human intervention.
-RL helps this process by allowing the agent to adapt to new device properties without the need for manual update of collection strategies.
-
-\subsection{Q-learning}
-
-Central to Q-learning, a widely used RL approach, is the action-value function $Q(S_t, A_t)$, which is the expected cumulative reward of taking action $A_t$ in state $S_t$.
-The agent approximates the optimal policy by updating the action-value function based on the reward that it receives during training\cite{suttonReinforcementLearningIntroduction2020}:
-
-\begin{equation}
-Q^{new}(S_t, A_t) \leftarrow Q(S_t, A_t) + \alpha \left[ R_{t+1} + \gamma \max_{a} Q(S_{t+1}, a) - Q(S_t, A_t) \right]
-\end{equation}
-
-where $\alpha$ is the learning rate, $\gamma$ is the discount factor, and $R_{t+1}$ and $S_{t+1}$ are the reward and state after taking action $A_t$ in state $S_{t}$, respectively.
-
-The agent's state consists of time of day $T$, battery power level $p$, and queued message count $m$.
-$T$ (discretized into 48 bins) informs the power generation likelihood, $p$ was discretized to 100 levels, and $m$ was discretized to 5 levels to enforce a transmission check every 2.5 hours to prevent data loss.
-This resulted in $T\times p \times m = 24,000$ possible states.
-The agent action space contains three discrete actions: $\mathcal{A}=\{a_0, a_1, a_2\}$, where $a_0=$delay data collection, $a_1=$collect and store data, and $a_2=$collect and transmit data, turning the action-value function into a $24000 \times 3$ table.
-
-A tabular representation for $Q(S_t, A_t)$ was used as it turns the model inference into a table lookup with $O(1)$ complexity.
-Deep Q-learning alternatives, which use a neural network to approximate $Q(S_t, A_t)$ require additional on-device computation and were not considered.
-A tabular representation does require larger memory or storage, but can be partially updated from the cloud to reflect new conditions.
-
-The goal of the agent is to maximize the number of messages sent to the cloud and the power level of the device.
-It is not trivial to determine the reward values that achieve this goal.
-We consider the reward values to be hyperparameters and used a Bayesian optimization framework from the python \verb|scikit-optimize| package to tune them.
-The hyperparameters (Table~\ref{tab:reward_parameters}) show that the most negative reward is for delaying collection. 
-The reward for transmitting data also collects the reward per message, which only becomes attractive if there are messages in the queue.
+\section{Simple Battery Drain Baseline}
+An initial experiment to verify that an agent is capable of learning involved a simple battery drain test.
+The battery was fully charged and the agent was run for as long as needed to discharge the battery while trying to maximize message collection/transmission.
+The agent was compared to a simple baseline static stragy of alternating transmission and collection.
+The baseline strategies are as follows:
+\begin{itemize}
+  \item Baseline 0: Take the Transmission action every time
+  \item Baseline 1: Collect once, transmit once
+  \item Baseline 2: Collect twice, transmit once
+  \item Baseline 3: Collect three times, transmit once
+  \item Baseline 4: Collect four times, transmit once
+\end{itemize}
 
@@
   \centering
-  \caption{Reward function parameters extracted during hyperparameter tuning.}
-  \begin{tabular}{l p{8cm} c}
+  \caption{Comparison of Baseline Models and RL Agent}
+  \begin{tabular}{lcccccc}
       \toprule
-      \textbf{Parameter} & \textbf{Description} & \textbf{Value} \\
+      & \textbf{Baseline 0} & \textbf{Baseline 1} & \textbf{Baseline 2} & \textbf{Baseline 3} & \textbf{Baseline 4} & \textbf{Agent} \\
       \midrule
-      reward\_power\_loss & reward for full discharge of battery & 0.00  \\ 
-      reward\_power\_multiplier & reward per power level at every time step & 0.0001  \\ 
-      reward\_action\_0 & reward for delaying collection & -1.716 \\ 
-      reward\_action\_1 & reward for collecting and storing data & -0.600 \\ 
-      reward\_action\_2 & reward for collecting and transmitting data & -1.485 \\ 
-      reward\_message\_count & reward for each message at transmission & 0.841 \\ 
+      Collection + Transmission Action & 205 & 117 & 81  & 62  & 50  & 54  \\
+      Collection Action                 & 0   & 117 & 162 & 186 & 200 & 198 \\
+      Total Messages                     & 205 & 234 & 243 & 248 & 250 & 252 \\
       \bottomrule
   \end{tabular}
-  \label{tab:reward_parameters}
+  \label{tab:batter_drain_result}
 \end{table}
 
-The agent was trained in a simulated environment with the power consumption and generation numbers collected from the prototype hardware.
-At each step, the environment updates the battery level, time of day, and queued messages based on the agent's action.
-The battery (discretized into 100 levels) drains 5.0 mAh per 30 minutes in sleep mode, while transmission and data collection consume 1.6 mAh and 0.1 mAh, respectively.
-
-Solar gain data from 17 June days in Edmonton was averaged into sunny and cloudy profiles, with an 80\% probability of a cloudy day. 
-This constraint limited power availability, forcing the agent to optimize message transmission. 
-The training process was repeated four times with different seeds for each hyperparameter search iteration. 
-Once the optimal hyperparameters were found, the action-value function was exported as an h-file for firmware deployment. 
-This automated approach reduced manual effort and effectively optimized reward values.
-
-\subsection{Hardware}
-The hardware prototype is shown in Figure \ref{fig:prototype} and consists of an Arduino microcontroller, a GPS device, a solar array, a power distribution system, and a battery.
-The detailed specifications are shown in Table \ref{tab:hardware_inventory}. 
-Prototype testing revealed high sleep power consumption (9.9 mA) caused by always-on LEDs and integrated components. 
-A custom PCB with full component isolation is planned to reduce sleep current to $\mu A$ levels.
-An unexpected benefit of high power draw was a faster development cycle, as battery depletion occurred in days rather than weeks.
-
-\begin{table}[h]
-  \centering
-  \caption{Hardware list for the cattle tracker prototype}
-  \begin{tabular}{l p{6.5cm} c}
-      \toprule
-      \textbf{Hardware} & \textbf{Description} & \textbf{Serial Number} \\
-      \midrule
-      Arduino MKR WiFi 1010 & WiFi-enabled microcontroller board & ABX00023 \\
-      Arduino MKR GPS Shield & Integrated GPS module & ASX00017 \\
-      DFRobot Solar Power Manager & Power distribution board & DFR0559 \\
-      Solar Panel & 75x100x2.9 mm panel, 1W output & TPX00181 \\
-      Sparkfun Fuel Gage & Power measurement sensor & MAX17043 \\
-      LiPo Battery & 3.7V, 2100mAh rated, 1300mAh eff. capacity & 2528 \\
-      \bottomrule
-  \end{tabular}
-  \label{tab:hardware_inventory}
-\end{table}
-
-\begin{figure}[h]
-  \centering
-  \begin{subfigure}[t]{0.48\textwidth}  % Align top
-      \centering
-      \includegraphics[height=6cm, keepaspectratio]{./figs/prototype.png}
-      \put(-170,160){\scriptsize \textbf{(A)}}  % Adjust position as needed
-      \label{fig:prototype_real}
-  \end{subfigure}
-  \hspace{2mm}  % Reduce space between figures
-  \begin{subfigure}[t]{0.48\textwidth}  % Align top
-      \centering
-      \includegraphics[height=6cm, keepaspectratio]{./figs/prototype_diagram.png}
-      \put(-215,160){\scriptsize \textbf{(B)}}  % Adjust position as needed
-      \label{fig:prototype_diagram}
-  \end{subfigure}
-  \caption{(A) Physical prototype with hardware components. (B) System architecture diagram illustrating data processing and power management.}
-  \label{fig:prototype}
-\end{figure}
-
-\subsection{Firmware}
-The prototype firmware, written in Arduino C++, handles sensor data collection, decision-making, and cloud transmission (see Figure~\ref{fig:prototype} B). 
-The main loop runs every 30 minutes, alternating between sleep mode and decision-making.
-At each decision point, the device reads the time, battery level, and queued messages, retrieves a recommended action from the policy table, and executes it.
-Due to memory constraints, only the highest-value action per state was stored, reducing the policy table size. 
-The policy was bit-shifted into an array of 32-bit integers for efficient storage in memory.
-
-\subsection{Cloud}
-The prototype transmits data to Azure IoT Hub, which integrates with IoT Central for dashboarding and Blob Storage for data retention (see Figure~\ref{fig:prototype} B). 
-IoT Central supports device registration, dashboard creation, and digital twins, enabling remote monitoring and state updates.
-The cloud facilitates two-way communication, allowing the device to receive policy updates for potential seasonal changes or location-specific solar gain variations.
-
-\section{Results}
-
-\subsection{Simulation Training Results}
-To evaluate the effectiveness of the reinforcement learning agent, we compare it to a baseline collection strategy that alternates between data collection and transmission, sending data once per hour.
-This represents a simple, fixed scheduling approach.
-Table~\ref{tab:agent_comparison} shows that the RL agent reduces transmissions, conserving power while increasing collection actions, leading to a higher message count (+4.3\%) and a more sustainable average power level (76 vs.
-34).
-The trained agent optimized collection-to-transmission behavior, performing 2.12 collection actions per transmission, compared to the 1:1 fixed schedule of the baseline. 
-This suggests the agent learned an adaptive strategy, balancing energy efficiency with message throughput.
-
-\begin{table}[h]
-  \centering
-  \caption{Comparison of baseline agent and RL agent}
-  \begin{tabular}{lcc}
-      \toprule
-      & Baseline Agent & RL Agent \\
-      \midrule
-      Transmission Actions & 480 & 308 \\
-      Collection Actions & 480 & 652 \\
-      Sleep Actions & 0 & 0 \\
-      Avg. Message Count & 756.5 & 789.5 \\
-      Avg. Power Level & 34 & 76 \\
-      \bottomrule
-  \end{tabular}
-  \label{tab:agent_comparison}
-\end{table}
-
-\subsection{Deployment Results}
-The trained agent was deployed on the prototype hardware in an indoor environment without power generation for a 9-day evaluation period, with occasional manual charging to ensure continuous operation.
-This test assessed the agent’s long-term decision-making performance in real hardware and validated its ability to operate under realistic power constraints.
-The results are shown in Figure~\ref{fig:prototype_result}.
-During the test, the hardware experienced several outages, the longest lasting 3.6 hours on 2024-10-07.  
-The agent maintained an average collection-to-transmission ratio of 2.18, which closely matches the simulation results.
-
-\begin{figure}[h]
-  \centering
-  \begin{subfigure}{0.48\textwidth}
-      \centering
-      \includegraphics[width=\linewidth]{figs/prototype_power_level.png}
-      \put(-195,138){\scriptsize \textbf{(A)}}  % Adjust position
-      \label{fig:prototype_power}
-  \end{subfigure}
-  \hfill
-  \begin{subfigure}{0.48\textwidth}
-      \centering
-      \includegraphics[width=\linewidth]{figs/battery_level_vs_collection_ratio.png}
-      \put(-195,138){\scriptsize \textbf{(B)}}  % Adjust position
-      \label{fig:battery_vs_collection}
-  \end{subfigure}
-  \caption{(A) Nonlinear power level discharge over a 9-day period. (B) Learned agent strategy adapting to varying power levels.}
-  \label{fig:prototype_result}
-\end{figure}
-
-Figure~\ref{fig:prototype_result}\subref{fig:battery_vs_collection}A illustrates the nonlinear battery discharge curve observed over the 9-day period.  
-The discharge rate during sleeping is relatively constant and the nonlinearity is a result of the mapping function of voltage to remaining capacity.
-While sleep mode consumption remains constant, the voltage-to-capacity mapping introduces nonlinearity, making it difficult for the agent to accurately predict future battery levels at low charge states.
-
-Figure~\ref{fig:prototype_result}\subref{fig:battery_vs_collection}B shows the agent's adaptive strategy under varying power levels.
-The agent alternates between collection and transmission at full charge, prioritizes storage with a 4:1 collection-to-transmission ratio at 30\% power, and transmits every collected message at low power which accelerates battery drain.
-Messages are stored in volatile memory so the queue is lost when the battery is fully discharged.
-The agent learned to transmit every time to prevent data loss at low power levels.
-This revealed a short-coming of the prototype and future iterations will store the queue in non-volatile memory to prevent data loss.
-
-\section{Conclusion}
-In this study, we developed a Q-learning-enabled cattle tracker that optimizes power consumption while ensuring efficient data collection and transmission. 
-The device dynamically adjusts its strategy based on energy availability, time of day, and message queue length, improving battery life and efficiency. 
-By integrating solar power with adaptive decision-making, the prototype offers a viable solution for long-term autonomous tracking. 
-Future work will focus on hardware optimizations, including low-power components and a custom PCB to enhance reliability and longevity. 
-Additionally, we aim to increase the message queue length, improve the reward function to account for the age of the messages, and incorporate non-volatile memory storage.
-
-\section*{Acknowledgements}
+The results are shown in Table \ref{tab:batter_drain_result}.
+The baseline is able to send the most messages if it collects four times and sends three times.
+This would seem like the most optimum strategy, but the static schedule gets into trouble when the battery runs out and the entire queue of messages is lost.
+The agent learns an optimal strategy of collecting four times and sending once except when the battery is low.
+In this case, the agent transmits more frequently to not lose the queue of messages.
 

### 2025-02-19 RL_nz/agent/agent.py

@@ -1,0 +1,12 @@
+from abc import ABC, abstractmethod
+
+class Agent(ABC):
+    @abstractmethod
+    def select_action(self, state):
+        pass
+    
+    def update_q_value(self, state, action, reward, next_state):
+        pass
+    
+    def decay_epsilon(self):
+        pass

### 2025-02-19 RL_nz/agent/baseline_agent.py

@@ -1,16 +1,14 @@
-class BaselineAgent:
-    def __init__(self, env):
+from agent.agent import Agent
+
+class BaselineAgent(Agent):
+    def __init__(self, config, env):
         self.N_ACTIONS = env.N_ACTIONS
-        self.current_action = 2  # Start with action 1
+        self.repeat_action1 = config.get("repeat_action1", 1)
+        self.period = self.repeat_action1 + 1
+        self.counter = 0
 
     def select_action(self, state):
-        action = self.current_action
-        # Alternate between action 1 and 2
-        self.current_action = 2 if self.current_action == 1 else 1
+        action = 1 if (self.counter % self.period) < self.repeat_action1 else 2
+        self.counter += 1
         return action
-    
-    def update_q_value(self, state, action, reward, next_state):
-        pass
-    
-    def decay_epsilon(self):
-        pass
+

### 2025-02-19 RL_nz/agent/tabular_agent.py

@@ -1,9 +1,10 @@
 import numpy as np
 import os
 import random
+from agent.agent import Agent
 from collections import deque
 
-class DQNAgent:
+class TabularAgent(Agent):
     def __init__(self, config, env):
         self.alpha = config['alpha']
         self.gamma = config['gamma']

### 2025-02-19 RL_nz/main.py

@@ -1,23 +1,0 @@
-import yaml
-from train import train, set_seed
-from agent.dqn_agent import DQNAgent
-from environment.custom_env import CustomEnv
-
-def main():
-    # Load configuration
-    with open('simple_battery_drain_optimum.yaml', 'r') as f:
-        config = yaml.safe_load(f)
-    
-    # Initialize environment and agent
-    env = CustomEnv(config['env'])
-    agent = DQNAgent(config['agent'], env)
-
-    # Start training
-    env.cloudy_chance = 1.0
-    set_seed(42)
-    message_count = train(env, agent, config['train'], plot_result_flag=True, result_prefix="simple_battery_drain_optimum_")
-    
-    
-
-if __name__ == "__main__":
-    main()

### 2025-02-19 RL_nz/simple_battery_drain_baseline.py

@@ -1,22 +1,37 @@
-import yaml
+from copy import deepcopy
 from train import train
 from agent.baseline_agent import BaselineAgent
+from agent.tabular_agent import TabularAgent
+from utils.yaml_helper import load_config
 from environment.custom_env import CustomEnv
 
-def main():
-    # Load configuration
-    with open('config_baseline.yaml', 'r') as f:
-        config = yaml.safe_load(f)
-    
-    # Initialize environment and agent
-    env = CustomEnv(config['env'])
-    baseline_agent = BaselineAgent(env)
-
-    # Try out baseline agent
-    env.cloudy_chance = 1.0
-    train(env, baseline_agent, config['train'], plot_result_flag=True, result_prefix="baseline_")
-    
-    
+# Function for the simple battery drain baseline experiment
 
 if __name__ == "__main__":
-    main()
+    # configure config files
+    config_list = []
+    prefix_list = []
+    agent_list = []
+    
+    agent_config = load_config('simple_battery_drain_optimum.yaml')
+    for k in range(5):
+        baseline_config = deepcopy(agent_config)
+        baseline_config['train']['num_episodes'] = 1
+        baseline_config['agent']['repeat_action1'] = k
+        config_list.append(baseline_config)
+        
+        prefix = f"simple_battery_drain_baseline_{k}_"
+        prefix_list.append(prefix)
+        agent_list.append(BaselineAgent)
+        
+    config_list.append(agent_config)
+    prefix_list.append("simple_battery_drain_optimum_")
+    agent_list.append(TabularAgent)
+    
+    for config, prefix, agent_class in zip(config_list, prefix_list, agent_list):
+        # Initialize environment and agent
+        env = CustomEnv(config['env'])
+        env.cloudy_chance = 1.0
+        
+        agent = agent_class(config['agent'], env)
+        train(env, agent, config['train'], plot_result_flag=True, result_prefix=prefix)

### 2025-02-19 RL_nz/simple_battery_drain_search.py

@@ -6,7 +6,7 @@ from skopt import gp_minimize
 from skopt.space import Real, Integer
 from skopt.utils import use_named_args
 from train import train, set_seed
-from agent.dqn_agent import DQNAgent
+from agent.tabular_agent import TabularAgent
 from environment.custom_env import CustomEnv
 
 
@@ -41,7 +41,7 @@ def main():
             
             # Initialize environment and agent
             env = CustomEnv(config['env'])
-            agent = DQNAgent(config['agent'], env)
+            agent = TabularAgent(config['agent'], env)
 
             # Start training
             env.cloudy_chance = 1.0

### 2025-02-19 RL_nz/solar_collection_baseline.py

@@ -22,7 +22,7 @@ def main():
     for i in range(1000):
         msg_list, pwr_list = train(env, baseline_agent, config['train'], plot_result_flag=False, result_prefix="baseline_solar_")
         message_list.append(msg_list[-1])
-        power_list.append(np.median(pwr_list[-1]))
+        power_list.append(np.median(pwr_list))
     
 
     print(f"Average messages sent: {np.median(message_list)}")
@@ -42,7 +42,7 @@ def main():
     axs[0].text(avg_power, 10, f'Median Power:\n {avg_power:.2f}', rotation=90, color='r')
     
     axs[1].hist(message_list)
-    # axs[1].set_title("Messages Sent")
+    axs[1].set_title("Messages Sent")
     # plot a red line showing the average messages sent
     avg_messages = np.median(message_list)
     axs[1].axvline(avg_messages, color='r', linestyle='dashed', linewidth=1)

### 2025-02-19 RL_nz/solar_connection_search.py

@@ -6,7 +6,7 @@ from skopt import gp_minimize
 from skopt.space import Real, Integer
 from skopt.utils import use_named_args
 from train import train, set_seed
-from agent.dqn_agent import DQNAgent
+from agent.tabular_agent import TabularAgent
 from environment.custom_env import CustomEnv
 
 
@@ -42,7 +42,7 @@ def main():
             
             # Initialize environment and agent
             env = CustomEnv(config['env'])
-            agent = DQNAgent(config['agent'], env)
+            agent = TabularAgent(config['agent'], env)
 
             # Start training
             message_count_list, power_list = train(env, agent, config['train'], plot_result_flag=False)

### 2025-02-19 RL_nz/solar_main_train.py

@@ -5,7 +5,7 @@ import matplotlib.pyplot as plt
 from datetime import date
 
 from train import train, set_seed
-from agent.dqn_agent import DQNAgent
+from agent.tabular_agent import TabularAgent
 from environment.custom_env import CustomEnv
 
 def main():
@@ -15,7 +15,7 @@ def main():
     
     # Initialize environment and agent
     env = CustomEnv(config['env'])
-    agent = DQNAgent(config['agent'], env)
+    agent = TabularAgent(config['agent'], env)
 
     # Start training
     env.cloudy_chance = 0.8

### 2025-02-20 RL_nz/agent/tabular_agent.py
Make the tabular_agent 10% faster

@@ -1,6 +1,5 @@
 import numpy as np
 import os
-import random
 from agent.agent import Agent
 from collections import deque
 
@@ -13,13 +12,13 @@ class TabularAgent(Agent):
         self.N_STATES = (env.N_POWER_LEVELS+1) * (env.N_TIME_INTERVALS+1) * env.MAX_MESSAGES
         self.N_ACTIONS = env.N_ACTIONS
         self.n_step = config.get('n_step', 3)  # Number of steps to look back
-
+        self.gamma_powers = self.gamma ** np.arange(self.n_step + 1)
         self.Q_matrix = np.random.uniform(low=-0.01, high=0.01, size=(self.N_STATES, self.N_ACTIONS))
         self.memory = deque(maxlen=self.n_step)
 
     def select_action(self, state):
-        if random.uniform(0, 1) < self.epsilon:
-            return random.randint(0, self.N_ACTIONS - 1)
+        if np.random.rand() < self.epsilon:
+            return np.random.randint(0, self.N_ACTIONS - 1)
         else:
             return np.argmax(self.Q_matrix[state])
         
@@ -34,17 +33,18 @@ class TabularAgent(Agent):
     def update_q_values(self):
         # Calculate the n-step return
         G = 0
-        for i in range(len(self.memory)):
+        n = len(self.memory)
+        for i in range(n):
             state, action, reward, next_state, done = self.memory[i]
-            G += (self.gamma ** i) * reward
+            G += self.gamma_powers[i] * reward
 
         # If the episode is not done, bootstrapping from the Q-value of the next state
         if not self.memory[-1][4]:  # Check if the last stored step is not done
             next_state = self.memory[-1][3]
-            G += (self.gamma ** len(self.memory)) * np.max(self.Q_matrix[next_state])
+            G += self.gamma_powers[n] * np.max(self.Q_matrix[next_state])
 
         # Update Q-values for each state-action pair in the memory
-        for i in range(len(self.memory)):
+        for i in range(n):
             state, action, reward, next_state, done = self.memory[i]
             self.Q_matrix[state][action] += self.alpha * (G - self.Q_matrix[state][action])
             # Reduce the return by the discounted reward

### 2025-02-20 RL_nz/environment/custom_env.py

@@ -2,12 +2,12 @@ import numpy as np
 
 from environment.device import Device
 
+CLOUDY_DAY = [0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0.34, 0, 0, 0, 0.34, 0, 0.34,0.34, 5.19, 5.197, 2.310, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0]
+SUNNY_DAY = [0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 13.2, 23.8, 49.6, 68.6, 58.0, 75.81, 74.89, 89.46, 86.70, 105, 105, 105, 105, 105, 97.3, 58.7, 58.7, 25.8, 20.3, 14.9, 20.4, 3.95, 9.43, 9.4, 3.9,0, 0, 0, 0, 0, 0, 0, 0, 0]
+SOLAR_DATA = [CLOUDY_DAY, SUNNY_DAY]
+
 def get_solar_intensity(time_step: int, current_day: int = 0):
-    cloudy_day = [0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0.34, 0, 0, 0, 0.34, 0, 0.34,0.34, 5.19, 5.197, 2.310, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0]
-    sunny_day = [0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 13.2, 23.8, 49.6, 68.6, 58.0, 75.81, 74.89, 89.46, 86.70, 105, 105, 105, 105, 105, 97.3, 58.7, 58.7, 25.8, 20.3, 14.9, 20.4, 3.95, 9.43, 9.4, 3.9,0, 0, 0, 0, 0, 0, 0, 0, 0]
-    
-    day_array = [cloudy_day, sunny_day]
-    return day_array[current_day][time_step]
+    return SOLAR_DATA[current_day][time_step]
     
 
 class CustomEnv:
@@ -24,6 +24,9 @@ class CustomEnv:
         self.current_day = 0
         self.cloudy_chance = config.get('cloudy_chance', 0.8)
         
[…]

### 2025-02-20 RL_nz/environment/device.py

@@ -24,7 +24,10 @@ class Device:
         self._rounding_factor = rounding_factor
         self._message_queue_length = 0
         self._max_message_queue_length = 5
-        pass
+        
+        # precomputed values
+        self._idle_collect = power_idle + power_collect
+        self._idle_transmit = power_idle + power_transmit
 
     
     def get_power_level_float(self) -> float:
@@ -42,22 +45,23 @@ class Device:
     def take_action(self, action: int) -> None:
         """ Takes an action and updates the power level accordingly. """
         messages_send = 0
+        power_current = self._power_current
         if action == 0:
-            self._power_current -= self._power_idle
+            power_current -= self._power_idle
         elif action == 1:
-            power_consumed = self._power_idle + self._power_collect
-            if self._power_current >= power_consumed:
+            power_consumed = self._idle_collect
+            if power_current >= power_consumed:
                 self._message_queue_length = min(self._message_queue_length+1, self._max_message_queue_length)
-            self._power_current -= power_consumed
+            power_current -= power_consumed
         elif action == 2:
-            power_consumed = self._power_idle + self._power_transmit
-            if self._power_current >= power_consumed:
+            power_consumed = self._idle_transmit
+            if power_current >= power_consumed:
                 messages_send = min(self._message_queue_length + 1, self._max_message_queue_length)
                 self._message_queue_length = 0
-            self._power_current -= power_consumed
+            power_current -= power_consumed
         else:
             raise ValueError("Invalid action")
-        self._power_current = max(0, self._power_current)
+        self._power_current = max(0, power_current)
         return messages_send
     
     def set_power_percentage(self, power_level_percent: float) -> None:

### 2025-02-20 paper/appendix.tex
Update to the agent to make it faster
- added comparissons to other types of baselines

@@
 
+\section{Solar Baseline}
+\begin{table}[h!]
+  \centering
+  \begin{tabular}{lcccccc}
+  \hline
+   & Baseline 0 & Baseline 1 & Baseline 2 & Baseline 3 & Baseline 4 & Agent \\
+  \hline
+  Transmission Actions & 960 & 480 & 320 & 240 & 190 & 308 \\
+  Collection Actions   & 0   & 480 & 640 & 720 & 770 & 652 \\
+  Sleep Actions        & 0   & 0   & 0   & 0   & 0   & 0   \\
+  Avg. Message Count   & 709 & 766.5 & 769 & 782 & 805 & 790 \\
+  Avg. Power Level     & 45  & 52  & 55  & 55  & 57  & 55  \\
+  \hline
+  \end{tabular}
+  \caption{Comparison of Baselines and Agent Performance}
+  \label{table:baselines_vs_agent}
+  \end{table}
+
 % All references should be stored in the file "references.bib".

### 2025-02-20 RL_nz/agent/tabular_agent.py

@@ -3,6 +3,18 @@ import os
 from agent.agent import Agent
 from collections import deque
 
+from numba import njit
+
+@njit
+def numba_argmax(arr):
+    max_val = arr[0]
+    max_index = 0
+    for i in range(1, arr.shape[0]):
+        if arr[i] > max_val:
+            max_val = arr[i]
+            max_index = i
+    return max_index
+
 class TabularAgent(Agent):
     def __init__(self, config, env):
         self.alpha = config['alpha']
@@ -20,7 +32,7 @@ class TabularAgent(Agent):
         if np.random.rand() < self.epsilon:
             return np.random.randint(0, self.N_ACTIONS - 1)
         else:
-            return np.argmax(self.Q_matrix[state])
+            return numba_argmax(self.Q_matrix[state])
         
     def store_step(self, state, action, reward, next_state, done):
         self.memory.append((state, action, reward, next_state, done))

### 2025-02-20 RL_nz/environment/device.py

@@ -1,4 +1,3 @@
-
 class Device:
     _power_current: float
     _power_max: float
@@ -75,8 +74,9 @@ class Device:
     
     def add_power(self, power: float) -> None:
         """ Adds power to the current power level """
-        self._power_current += power
-        self._power_current = min(self._power_current, self._power_max)
+        power_current = self._power_current
+        power_current += power
+        self._power_current = min(power_current, self._power_max)
         
     
     

### 2025-02-20 RL_nz/solar_collection_baseline.py

@@ -1,59 +1,78 @@
 import matplotlib.pyplot as plt
 import numpy as np
 import yaml
+from copy import deepcopy
 from train import train, set_seed
 from agent.baseline_agent import BaselineAgent
+from agent.tabular_agent import TabularAgent
 from environment.custom_env import CustomEnv
-
-def main():
-    # Load configuration
-    with open('solar_config.yaml', 'r') as f:
-        config = yaml.safe_load(f)
-    config['train']['num_episodes'] = 1
-    
-    # Initialize environment and agent
-    env = CustomEnv(config['env'])
-    baseline_agent = BaselineAgent(env)
-
-    # Try out baseline agent
-    env.cloudy_chance = 0.8
-    message_list = []
-    power_list = []
-    for i in range(1000):
-        msg_list, pwr_list = train(env, baseline_agent, config['train'], plot_result_flag=False, result_prefix="baseline_solar_")
-        message_list.append(msg_list[-1])
-        power_list.append(np.median(pwr_list))
-    
-
-    print(f"Average messages sent: {np.median(message_list)}")
-    print(f"Average power: {np.median(power_list)}")
-    
-    # Show the results
-    
-    # show a side by side plot with the power and messages
-    fig, axs = plt.subplots(2, figsize=(7, 8))
-    fig.suptitle("Baseline Agent")
-    axs[0].hist(power_list)
-    axs[0].set_title("Power")
-    # plot a red line showing the average power
-    avg_power = np.median(power_list)
-    axs[0].axvline(avg_power, color='r', linestyle='dashed', linewidth=1)
-    # add text to the plot showing the average power
-    axs[0].text(avg_power, 10, f'Median Power:\n {avg_power:.2f}', rotation=90, color='r')
-    
-    axs[1].hist(message_list)
-    axs[1].set_title("Messages Sent")
-    # plot a red line showing the average messages sent
-    avg_messages = np.median(message_list)
-    axs[1].axvline(avg_messages, color='r', linestyle='dashed', linewidth=1)
-    # add text to the plot showing the average power
-    axs[1].text(avg_messages, 10, f'Median Msg:\n {avg_messages:.2f}', rotation=90, color='r')
-    axs[1].set_ylim(0, 300)
-    
-    axs[0].set_xlabel("Power [%]")
-    axs[1].set_xlabel("Messages Sent")
-    plt.show()
+from utils.yaml_helper import load_config
 
 if __name__ == "__main__":
[…]

### 2025-02-20 RL_nz/solar_collection_search.py

@@ -17,13 +17,13 @@ def main():
 
     # Define the search space
     space = [
-        # Integer(3000, 6000,  name='num_episodes'),
-        Real(-15.0, 0.0, name='reward_power_loss'),
-        Real(0.0001, 0.01, name='reward_power_multiplier'),
-        Real(-2.0, 2.0, name='reward_action_0'),
-        Real(-2.0, 2.0, name='reward_action_1'),
-        Real(-2.0, 2.0, name='reward_action_2'),
-        Real(0.1, 3.0, name='reward_message_count')
+        Integer(1000, 6000,  name='num_episodes'),
+        # Real(-15.0, 0.0, name='reward_power_loss'),
+        Real(-0.5, 0.5, name='reward_power_multiplier'),
+        Real(-3.0, 3.0, name='reward_action_0'),
+        Real(-3.0, 3.0, name='reward_action_1'),
+        Real(-3.0, 3.0, name='reward_action_2'),
+        Real(-3.0, 3.0, name='reward_message_count')
     ]
     
     @use_named_args(space)

### 2025-02-21 RL_nz/train.py
Modify the train results to return the median power list

@@ -1,7 +1,7 @@
 import matplotlib.pyplot as plt
 import numpy as np
 import random
-
+from agent.agent import Agent
 from environment.custom_env import CustomEnv
 from utils.plot import plot_results
 from typing import List, Tuple
@@ -37,10 +37,15 @@ def save_results(filename, data):
         writer.writerow(["Episode", "Value"])
         writer.writerows(enumerate(data, 1))
 
-def train(env: CustomEnv, agent, train_config, plot_result_flag:bool=True, result_prefix:str="") -> Tuple[List[int], List[float]]:
+def train(env: CustomEnv, 
+          agent: Agent, 
+          train_config:dict, 
+          plot_result_flag:bool=True, 
+          result_prefix:str="") -> Tuple[List[int], List[float]]:
     num_episodes = train_config['num_episodes']
     reward_list = []
     legitimate_messages_list = []
+    median_power_list = []
 
     for episode in range(num_episodes):
         env.init_state(power_level=100, time=0, message_count=0)
@@ -66,6 +71,7 @@ def train(env: CustomEnv, agent, train_config, plot_result_flag:bool=True, resul
         agent.decay_epsilon()
         reward_list.append(total_reward)
         legitimate_messages_list.append(legitimate_messages_sent_this_episode)
+        median_power_list.append(np.median(power_list))
         if plot_result_flag:
             print(f'Episode {episode+1}: Total Reward = {total_reward}, Legitimate Messages Sent = {legitimate_messages_sent_this_episode}')
                 
@@ -85,4 +91,4 @@ def train(env: CustomEnv, agent, train_config, plot_result_flag:bool=True, resul
         save_results(f'results/{result_prefix}power_list.csv', power_list)
         save_results(f'results/{result_prefix}power_list.csv', action_list)
         
-    return legitimate_messages_list, power_list
+    return legitimate_messages_list, median_power_list

### 2025-02-23 RL_nz/environment/base_device.py
Implement gym environment

@@ -1,0 +1,27 @@
+from abc import ABC, abstractmethod
+
+class BaseDevice(ABC):
+    @abstractmethod
+    def set_power_percentage(self, power_level_percent: float) -> None:
+        """Set the current power level as a percentage of maximum power."""
+        pass
+
+    @abstractmethod
+    def calculate_rounded_power_level(self) -> int:
+        """Return the current power level, rounded to the nearest integer."""
+        pass
+
+    @abstractmethod
+    def set_message_queue_length(self, message_count: int) -> None:
+        """Set the current message queue length."""
+        pass
+
+    @abstractmethod
+    def add_power(self, power: float) -> None:
+        """Add generated power to the current power level."""
+        pass
+
+    @abstractmethod
+    def take_action(self, action: int) -> int:
+        """Apply an action, update the internal state, and return the number of messages sent."""
+        pass

### 2025-02-23 RL_nz/environment/device.py

@@ -1,4 +1,6 @@
-class Device:
+from environment.base_device import BaseDevice
+
+class Device(BaseDevice):
     _power_current: float
     _power_max: float
     _power_idle: float
@@ -21,6 +23,8 @@ class Device:
         self._power_transmit = power_transmit
         self._power_collect = power_collect
         self._rounding_factor = rounding_factor
+        
+        # Message queue parameters
         self._message_queue_length = 0
         self._max_message_queue_length = 5
         
@@ -70,7 +74,6 @@ class Device:
     def set_message_queue_length(self, message_count: int) -> None:
         """ Sets the message queue length """
         self._message_queue_length = message_count
-
     
     def add_power(self, power: float) -> None:
         """ Adds power to the current power level """

### 2025-02-23 RL_nz/environment/gym_environment.py

@@ -1,0 +1,129 @@
+import gymnasium as gym
+
+from gymnasium import spaces
+import numpy as np
+
+CLOUDY_DAY = [0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0.34, 0, 0, 0, 0.34, 0, 0.34,0.34, 5.19, 5.197, 2.310, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0]
+SUNNY_DAY = [0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 13.2, 23.8, 49.6, 68.6, 58.0, 75.81, 74.89, 89.46, 86.70, 105, 105, 105, 105, 105, 97.3, 58.7, 58.7, 25.8, 20.3, 14.9, 20.4, 3.95, 9.43, 9.4, 3.9,0, 0, 0, 0, 0, 0, 0, 0, 0]
+SOLAR_DATA = [CLOUDY_DAY, SUNNY_DAY]
+
+def get_solar_intensity(time_step: int, current_day: int = 0):
+    return SOLAR_DATA[current_day][time_step]
+
+class CustomGymEnv(gym.Env):
+    """
+    A Gymnasium environment that simulates device behavior under variable solar power conditions.
+    The environment uses an injected device (an instance of a class inheriting from BaseDevice)
+    to handle power and action logic.
+    """
+    def __init__(self, config, device):
+        super().__init__()
+        
+        # Environment configuration
+        self.N_TIME_INTERVALS = config['time_intervals']
+        self.N_DAYS = config['n_days']
+        self.action_list = config['action_list']
+        self.N_ACTIONS = len(self.action_list)
+        self.N_POWER_LEVELS = config['power_levels']
+        self.MAX_MESSAGES = config['max_messages']
+        self.cloudy_chance = config.get('cloudy_chance', 0.8)
+        
+        # Define Gymnasium action and observation spaces
+        self.action_space = spaces.Discrete(self.N_ACTIONS)
+        self.observation_space = spaces.Discrete(self.N_TIME_INTERVALS * self.MAX_MESSAGES * self.N_POWER_LEVELS)
+        self._device = device
+        self.missed_message_penalty = config.get('missed_message_penalty', 1.5)
+        
+        
+        # Internal counters
+        self._time = 0
[…]

### 2025-02-23 RL_nz/solar_main_train.py

@@ -4,53 +4,119 @@ import numpy as np
 import matplotlib.pyplot as plt
 from datetime import date
 
-from train import train, set_seed
-from agent.tabular_agent import TabularAgent
-from environment.custom_env import CustomEnv
+# from train import train, set_seed
 
-def main():
-    # Load configuration
-    with open('solar_config.yaml', 'r') as f:
-        config = yaml.safe_load(f)
+import gymnasium as gym
+from agent.tabular_agent import TabularAgent
+from environment.device import Device
+from environment.gym_environment import CustomGymEnv
+from utils.yaml_helper import load_config
+
+if __name__=='__main__':
+    config = load_config('solar_config.yaml')
+    device = Device()
+    env = CustomGymEnv(config['env'], device)
     
     # Initialize environment and agent
-    env = CustomEnv(config['env'])
     agent = TabularAgent(config['agent'], env)
 
-    # Start training
-    env.cloudy_chance = 0.8
-    message_list, power_list = train(env, agent, config['train'], plot_result_flag=True, result_prefix="solar_collection_optimum_")
-    
-    print(f"Average messages sent: {np.median(message_list[-1000:])}")
-    print(f"Average power: {np.median(power_list[-1000:])}")
-    
-    # Save the Q-matrix and add the date to the filename
-    filename = f"./results/models/{date.today().strftime('%Y-%m-%d')}_solar_collection_optimum_q_matrix.npy"
-    agent.save_q_matrix(filename)
-    
-        # show a side by side plot with the power and messages
-    fig, axs = plt.subplots(2, figsize=(7, 8))
-    fig.suptitle("RL Agent")
-    axs[0].hist(power_list[-1000:])
-    axs[0].set_title("Power")
-    
-    avg_power = np.median(power_list[-1000:])
-    axs[0].axvline(avg_power, color='r', linestyle='dashed', linewidth=1)
-    axs[0].text(avg_power, 10, f'Median Power:\n {avg_power:.2f}', rotation=90, color='r')
-    
-    axs[1].hist(message_list[-1000:])
 
-    avg_messages = np.median(message_list[-1000:])
-    axs[1].axvline(avg_messages, color='r', linestyle='dashed', linewidth=1)
-    axs[1].text(avg_messages, 10, f'Median Msg:\n {avg_messages:.2f}', rotation=90, color='r')
+    for episode in range(config['train']['num_episodes']):
+        observation, info = env.reset()
[…]

### 2025-03-02 RL_nz/agent/tabular_agent.py
Fix bug introduced during refactoring, which prevents explroation of the sending state

@@ -48,12 +48,12 @@ class TabularAgent(Agent):
 
     def discretize_state(self, state_vector):
         # Turn the state vector of [power, time, queue] into an integer index
-        return int(state_vector[0]) * (self.num_time_bins * self.num_queue_bins) + int(state_vector[1]) * self.num_queue_bins + int(state_vector[2])
+        return int(state_vector[0]) * (self.num_time_bins * self.num_queue_bins) + int(state_vector[1] * self.num_queue_bins) + int(state_vector[2])
 
     def select_action(self, state):
         state = self.process_state(state)
         if np.random.rand() < self.epsilon:
-            return np.random.randint(0, self.N_ACTIONS - 1)
+            return np.random.randint(0, self.N_ACTIONS)
         else:
             return numba_argmax(self.Q_matrix[state])
         

### 2025-03-03 RL_nz/environment/gym_environment.py
Update the environment to allow for two types of rewards:
- shaped rewards that give rewards for actions
- general rewards that reward outcomes

@@ -1,5 +1,6 @@
 import gymnasium as gym
 
+from environment.base_device import BaseDevice
 from gymnasium import spaces
 import numpy as np
 
@@ -19,7 +20,8 @@ class CustomGymEnv(gym.Env):
     def __init__(self, 
                  config, 
                  device,
-                 normalize_state: bool = False):
+                 normalize_state: bool = False,
+                 use_reward_shaping: bool = False):
         """Initializes the environment with the given configuration and device."""
         super().__init__()
         
@@ -32,15 +34,24 @@ class CustomGymEnv(gym.Env):
         self.N_ACTIONS = len(self.action_list)
         self.cloudy_chance = config.get('cloudy_chance', 0.8)
         self.normalize_state = normalize_state
+        self.use_reward_shaping = use_reward_shaping
         
         # Define Gymnasium action and observation spaces
         self.action_space = spaces.Discrete(self.N_ACTIONS)
         self.observation_space = spaces.Box(low=0.0, high=1.0, shape=(3,), dtype=np.float32)
-        self._device = device
+        self._device: BaseDevice = device
+
+        # Reward parameters - Messsage Based
         self.message_reward = config.get('message_reward', 1.0)
         self.missed_message_penalty = config.get('missed_message_penalty', -1.5)
         self.power_reward = config.get('power_reward', 0.0001)
-        
+
+        # Reward shaping parameters
+        self.reward_power_multiplier: float = config.get('reward_power_multiplier', 0.0001)
+        self.reward_action_0: float = config.get('reward_action_0', -2.0)
+        self.reward_action_1: float = config.get('reward_action_1', -0.7353431237036132)
+        self.reward_action_2: float = config.get('reward_action_2', -0.5885436726404254)
+        self.reward_message_count: float = config.get('reward_message_count', 0.8863782281442544)
         
         # Internal counters
         self._time = 0
@@ -65,6 +76,18 @@ class CustomGymEnv(gym.Env):
         observation = self.get_state_vector()
         return observation, {}
 
+    def _get_shaped_reward(self, action:int, messages_sent:int):
+        action_reward = 0
+        if action == 0:
+            action_reward = self.reward_action_0
+        elif action == 1:
+            action_reward = self.reward_action_1
+        elif action == 2:
+            action_reward = self.reward_action_2
+        else:
+            raise ValueError(f"Invalid action: {action}")
+        return action_reward + messages_sent * self.reward_message_count + self.get_power() * self.reward_power_multiplier
+
 
     def step(self, action):
[…]

### 2025-03-04 paper/appendix.tex
Update the appendix with additional information around simple battery drain

@@
 An initial experiment to verify that an agent is capable of learning involved a simple battery drain test.
+The purpose was to check the agents' ability to learna long-term strategy.
+It is simple enough to take early rewards, but in a battery drain test, the agent must learn to balance short-term rewards with long-term goals and conserve energy.
 The battery was fully charged and the agent was run for as long as needed to discharge the battery while trying to maximize message collection/transmission.
+The power was discretized into 400 power levels, the day was divided into 48 time steps, and the maxmimum message queue size was 5 messages and ran for 6 days until the battery was drained.
+
+\subsection{Agent Tuning}
+
+There were two optimizatiosn performed to tune the agent.
+The first optimization involved the hyper paper tuning of alpha, gamma, epsilon, epsilon decay, $n\_steps$, episode count with rewards of 1.0 per send message and \-1.5 per lost message and 0.0001 of power multiplier.
+The second optimization was a reward shaping optimization and tuned rewards for performing each action, the power multiplier, and the reward for sending a message.
+
+The hyperparameter tuning resulted with various parameters where the agent reached a maximum message count of 230.
+It would appear that the agent was not able to learn a long-term strategy and was only able to learn a short-term strategy for reward maximization.
+
+The reward shaping optimization resulted in an agent with a maximum message count of 253.
+The agent was able to learn a long-term strategy and was able to balance short-term rewards with long-term goals.
+The agent appears to need intermediate rewards to learn a long-term strategy.
+The followin parameters were found: power reward =0.000, Action0=-1.17324902,Action1=-0.463915495, Action2=-2, PerMessage=	0.602358605
+
+\subsection{Battery Drain Test Results}
+
 The agent was compared to a simple baseline static stragy of alternating transmission and collection.
@@
 The results are shown in Table \ref{tab:batter_drain_result}.
-The baseline is able to send the most messages if it collects four times and sends three times.
+The baseline is able to send the most messages if it collects four times and sends once.
 This would seem like the most optimum strategy, but the static schedule gets into trouble when the battery runs out and the entire queue of messages is lost.

### 2025-03-04 RL_nz/simple_battery_drain_baseline.py
Add a tuning function for hyper parameters

@@ -14,7 +14,7 @@ if __name__ == "__main__":
     agent_list = []
     
     agent_config = load_config('simple_battery_drain_optimum.yaml')
-    for k in range(5):
+    for k in range(4):
         baseline_config = deepcopy(agent_config)
         baseline_config['train']['num_episodes'] = 1
         baseline_config['agent']['repeat_action1'] = k

### 2025-03-04 RL_nz/simple_battery_drain_hp_tune.py

@@ -1,0 +1,75 @@
+import numpy as np
+import time
+import yaml
+
+from skopt import gp_minimize
+from skopt.space import Real, Integer
+from skopt.utils import use_named_args
+
+from train import train_gymnasium, set_seed  # Assume set_seed is defined
+from agent.tabular_agent import TabularAgent
+from environment.gym_environment import CustomGymEnv
+from environment.device import Device
+from utils.yaml_helper import load_config
+
+def main():
+    # Load base configuration from YAML file.
+    config = load_config('simple_battery_drain_optimum.yaml')
+
+    # Define the search space for hyperparameters.
+    search_space = [
+        Integer(1000, 5000, name='num_episodes'),
+        Real(0.001, 0.2, name='alpha'),
+        Real(0.01, 1.0, name='epsilon'),
+        Real(0.90, 0.9999, name='epsilon_decay'),
+        Real(0.9, 0.9999, name='gamma'),
+        # Integer(1, 50, name='n_step')
+    ]
+    
+    @use_named_args(search_space)
+    def objective(**params):
+        start_time = time.time()
+        message_counts = []
+        n_runs = 3  # Use multiple runs to reduce variance.
+        for i in range(n_runs):
+            # Set the seed for reproducibility.
+            set_seed(42 + i)
+            
+            # Update the training configuration with the current hyperparameters.
+            config['train']['num_episodes'] = params['num_episodes']
+            config['agent']['alpha'] = params['alpha']
+            config['agent']['epsilon'] = params['epsilon']
+            config['agent']['epsilon_decay'] = params['epsilon_decay']
+            config['agent']['gamma'] = params['gamma']
+            # config['agent']['n_step'] = int(params['n_step'])
+            
+            # Create the device and Gymnasium environment.
+            device = Device(power_max=config['env']['max_power'],
+                            rounding_factor=(100 / config['env']['power_levels']))
+            env = CustomGymEnv(config['env'], device, normalize_state=False, use_reward_shaping=True)
+            env.cloudy_chance = 1.0  # Fix the weather condition for this experiment.
+            
+            # Initialize the TabularAgent with the current config.
+            agent = TabularAgent(config['agent'], env)
+            
+            # Run training using the gymnasium training function.
+            message_count_list, _ = train_gymnasium(env, agent, config['train'], plot_result_flag=False)
+            # Record the number of messages from the last episode.
+            message_counts.append(np.median(message_count_list[-100:]))
+        
[…]

### 2025-03-16 RL_nz/simple_battery_drain_baseline.py
Add an additional evaluation step

@@ -1,3 +1,4 @@
+import matplotlib.pyplot as plt
 from copy import deepcopy
 from train import train_gymnasium
 from agent.baseline_agent import BaselineAgent
@@ -14,15 +15,15 @@ if __name__ == "__main__":
     agent_list = []
     
     agent_config = load_config('simple_battery_drain_optimum.yaml')
-    for k in range(4):
-        baseline_config = deepcopy(agent_config)
-        baseline_config['train']['num_episodes'] = 1
-        baseline_config['agent']['repeat_action1'] = k
-        config_list.append(baseline_config)
+    # for k in range(4):
+    #     baseline_config = deepcopy(agent_config)
+    #     baseline_config['train']['num_episodes'] = 1
+    #     baseline_config['agent']['repeat_action1'] = k
+    #     config_list.append(baseline_config)
         
-        prefix = f"simple_battery_drain_baseline_{k}_"
-        prefix_list.append(prefix)
-        agent_list.append(BaselineAgent)
+    #     prefix = f"simple_battery_drain_baseline_{k}_"
+    #     prefix_list.append(prefix)
+    #     agent_list.append(BaselineAgent)
         
     config_list.append(agent_config)
     prefix_list.append("simple_battery_drain_optimum_")
@@ -34,10 +35,53 @@ if __name__ == "__main__":
             power_max=config['env']['max_power'],
             rounding_factor=(100 / config['env']['power_levels'])
         )
-        env = CustomGymEnv(config['env'], device, normalize_state=False, use_reward_shaping=True)
+        env = CustomGymEnv(config['env'], device, normalize_state=False, use_reward_shaping=False)
         env.cloudy_chance = 1.0
         
         agent = agent_class(config['agent'], env)
         train_gymnasium(env, agent, config['train'], plot_result_flag=True, verbose=True, result_prefix=prefix)
         
+    # Evaluate the agent
+    obs, info = env.reset()
+    episode_reward = 0.0
+    episode_messages = 0
+    episode_power = []
+    action_list = []
+    done = False
+    agent.epsilon = 0.0
+    
+    while not done:
+        # Agent selects an action based on current observation.
+        action = agent.select_action(obs)
+        # Environment executes the action.
+        obs_next, reward, terminated, truncated, info = env.step(action)
+        done = terminated or truncated
+        
+        # Update agent using the experience.
[…]

### 2025-03-18 RL_nz/train.py
Update the traing script

@@ -140,7 +140,7 @@ def train_gymnasium(env: CustomGymEnv,
             
             # Accumulate statistics.
             episode_reward += reward
-            episode_messages += info.get('messages_sent', 0)
+            episode_messages = info.get('messages_sent', 0)
             episode_power.append(info.get('device_power', 0))
             action_list.append(action)
             
@@ -151,7 +151,7 @@ def train_gymnasium(env: CustomGymEnv,
         agent.decay_epsilon()
         
         total_message_list.append(episode_messages)
-        power_list.extend(episode_power)
+        power_list = episode_power
         if verbose:
             print(f"Episode {episode+1:3d}: Total Reward = {episode_reward:7.3f}, "
                 f"Messages = {episode_messages}, Epsilon = {agent.epsilon:7.6f}")

### 2025-03-18 paper/appendix.tex
Fix typos in appendinx

@@
 There were two optimizatiosn performed to tune the agent.
-The first optimization involved the hyper paper tuning of alpha, gamma, epsilon, epsilon decay, $n\_steps$, episode count with rewards of 1.0 per send message and \-1.5 per lost message and 0.0001 of power multiplier.
+The first optimization involved the hyperparameter tuning of alpha, gamma, epsilon, epsilon decay, $n\_steps$, episode count with rewards of 1.0 per send message and \-1.5 per lost message and 0.0001 of power multiplier.
 The second optimization was a reward shaping optimization and tuned rewards for performing each action, the power multiplier, and the reward for sending a message.
@@
 The agent appears to need intermediate rewards to learn a long-term strategy.
-The followin parameters were found: power reward =0.000, Action0=-1.17324902,Action1=-0.463915495, Action2=-2, PerMessage=	0.602358605
+The following parameters were found: power reward =0.000, Action0=-1.17324902,Action1=-0.463915495, Action2=-2, PerMessage=	0.602358605
+
+It is interesting that the reward shaping produced a better result than pure message count.
+The reward shaping shows the agent which actions should be rewarded to produce a better message count.
+This is likely because the gradient of rewards did not provide enough information to the agent to learn a long-term strategy for the hyper parameter tuning case.
 

### 2025-03-27 paper/appendix.tex
add some notes to the appendix

@@
 An initial experiment to verify that an agent is capable of learning involved a simple battery drain test.
-The purpose was to check the agents' ability to learna long-term strategy.
+The purpose was to check the agents' ability to learn a long-term strategy.
 It is simple enough to take early rewards, but in a battery drain test, the agent must learn to balance short-term rewards with long-term goals and conserve energy.
 The battery was fully charged and the agent was run for as long as needed to discharge the battery while trying to maximize message collection/transmission.
-The power was discretized into 400 power levels, the day was divided into 48 time steps, and the maxmimum message queue size was 5 messages and ran for 6 days until the battery was drained.
+The power was discretized into 400 power levels, the day was divided into 48 time steps, and the maximum message queue size was 5 messages and ran for 6 days until the battery was drained.
 
@@
 
-The reward shaping optimization resulted in an agent with a maximum message count of 253.
-The agent was able to learn a long-term strategy and was able to balance short-term rewards with long-term goals.
+The reward shaping optimization resulted in an agent with a maximum message count of 253 or 252 depending on which seed is selected.
+The optimization found reward values that the agent used to come up with a strategy that happens to outperform all baselines.
 The agent appears to need intermediate rewards to learn a long-term strategy.
 The following parameters were found: power reward =0.000, Action0=-1.17324902,Action1=-0.463915495, Action2=-2, PerMessage=	0.602358605
+The agent is able to outperform the baseline because the baseline runs out of power with a full message queue and never gets to send the message.
 
@@
 This is likely because the gradient of rewards did not provide enough information to the agent to learn a long-term strategy for the hyper parameter tuning case.
+This methodology isn't amazing, because we pretty much guide the agent every step of the way by giving rewards for certain actions, rather than outcomes. 
+A different approach is likely necessary to ensure that the prcess is more robust.
 
